---
title: "Subgroups"
format: gfm
---

[Source](https://schochastics.github.io/R4SNA/descriptive/clustering.html)

```{r}
options(paged.print = FALSE)
```

```{r}
#| message: false

libraries <- list(
  "igraph", "networkdata", "blockmodeling",
  "netUtils", "ggraph", "patchwork"
)
invisible(lapply(libraries, library, character.only = TRUE))
```

# Cliques

A _clique_ is a set of nodes where every pair is directly connected, forming a complete _subgraph_. A _maximal clique_ cannot be extended by adding more nodes while retaining the property. It is the most strict definition of a cohesive subgroup, but is rarely encountered in real networks

```{r}
#| label: fig-clique-graph-blank
#| fig-cap: "The example network used to illustrate cliques and k-core decomposition."

ggraph(clique_graph, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(shape = 21, size = 8, fill = "grey25") +
  theme_void()
```

`max_cliques()` is only feasible for smaller networks.

```{r}
(cl <- max_cliques(clique_graph, min = 3))
```

```{r}
#| label: fig-clique-graph
#| fig-cap: "Maximal cliques of size 3 or more highlighted in the example network."

xy <- graphlayouts::layout_with_stress(clique_graph)

cl_df <- as.data.frame(do.call(
  "rbind",
  lapply(seq_along(cl), function(x) {
    cbind(xy[cl[[x]], ], x)
  })
))

ggraph(clique_graph, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(shape = 21, size = 8, fill = "grey25") +
  ggforce::geom_mark_hull(
    data = cl_df,
    aes(V1, V2, fill = as.factor(x), group = x),
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c(
      "#E69F00",
      "#000000",
      "#56B4E9",
      "#009E73",
      "#F0E442",
      "#0072B2",
      "#D55E00",
      "#CC79A7",
      "#666666"
    )
  ) +
  theme_void()
```

_K-core decomposition_ creates subgraphs in which each node has at least k neighbors. It is a relaxed version of a clique.

```{r}
(kcore <- coreness(clique_graph))
```

```{r}
#| label: fig-kcore
#| fig-cap: "K-core decomposition of the example network. Nested hulls show the 2-core, 3-core, and 4-core."

cl_df <- as.data.frame(do.call(
  "rbind",
  lapply(sort(unique(kcore))[c(2, 3, 4)], function(x) {
    cbind(xy[kcore >= x, ], x)
  })
))

ggraph(clique_graph, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(shape = 21, size = 8, fill = "grey25") +
  ggforce::geom_mark_hull(
    data = cl_df,
    aes(V1, V2, fill = as.factor(x), group = x),
    show.legend = FALSE
  ) +
  scale_fill_manual(values = c("red", "blue", "green")) +
  theme_void()
```

# Community Detection

A fuzzily defined concept, referring to groups of nodes which are more densely connected to each other than to the rest of the network. Many different algorithms exist. Most are based on "modularity maximization". Modularity compares the observed number of edges within groups to the expectation for forming random edges while maintaining the degree of each node.

Modularity:

$$
Q = \frac{1}{2m} \sum_{ij} \left[ A_{ij} - \frac{k_i k_j}{2m} \right] \delta(c_i, c_j)
$$

where $m$ is the total number of edges, $A_{i,j}$ is the adjacency matrix entry (1 if nodes $i$ and $j$ are connected, 0 otherwise), $k_i$ is the degree of node $i$, and $\delta(c_i, c_j)$ equals 1 if nodes $i$ and $j$ are assigned to the same community and 0 otherwise. The term $\frac{k_i k_j}{2m}$ is the expected number of edges between $i$ and $j$ under a null model that preserves degree. Modularity values typically range from around -0.5 to 1, where values close to zero indicate no more within-group edges than expected by chance, and higher values indicate stronger community structure.

The workflow of a cluster analysis is always the same, independent from the chosen method. We illustrate the workflow using the infamous karate club network, already encountered in Chapter 3. This network was collected by Wayne Zachary in the 1970s and it captures friendships among 34 members of a university karate club. During the study, a conflict between the club’s instructor and the administrator led to the club splitting into two factions, making it a classic benchmark for community detection.

```{r}
#| label: fig-karate-plot
#| fig-cap: "Zachary's karate club network."

ggraph(karate, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(shape = 21, size = 5, fill = "grey66") +
  theme_void() +
  coord_equal()
```

Using the Louvain method.

```{r}
clu <- cluster_louvain(karate)

(mem <- membership(clu))
```

```{r}
(com <- communities(clu))
```

Comparing different methods

```{r}
imc <- cluster_infomap(karate)
lec <- cluster_leading_eigen(karate)
loc <- cluster_louvain(karate)
sgc <- cluster_spinglass(karate)
wtc <- cluster_walktrap(karate)
scores <- c(
  infomap = modularity(karate, membership(imc)),
  eigen = modularity(karate, membership(lec)),
  louvain = modularity(karate, membership(loc)),
  spinglass = modularity(karate, membership(sgc)),
  walk = modularity(karate, membership(wtc))
)
scores
```

For networks up to around 100 nodes, the optimal partition can be calculated:

```{r}
optc <- cluster_optimal(karate)
modularity(karate, membership(optc))
```

```{r}
#| label: fig-karate-clu
#| fig-cap: "Karate club network colored by community membership (optimal partition)."

V(karate)$clu <- membership(optc)
ggraph(karate, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(
    shape = 21,
    size = 5,
    aes(fill = as.factor(clu)),
    show.legend = FALSE
  ) +
  theme_void() +
  coord_equal()
```

Note that when modularity is being maximized, smaller clusters can be merged together to form bigger clusters. In a ring of 50 clusters of size 5, the Louvain method merges multiple cliques resulting in an unintuitive clustering.

```{r}
#| label: fig-K50-blank
#| fig-cap: "A ring of 50 cliques of size 5, used to illustrate the resolution limit."
#| width: 8
#| height: 8
n1 <- 5
n2 <- 50
A <- matrix(1, n1, n1)
lst <- vector("list", n2)
lst <- lapply(lst, function(x) A)
AA <- Matrix::bdiag(lst)
for (i in 1:(n2 - 1)) {
  AA[i * n1, i * n1 + 1] <- AA[i * n1 + 1, i * n1] <- 1
}
AA[1, n1 * n2] <- AA[n1 * n2, 1] <- 1
K50 <- graph_from_adjacency_matrix(AA, "undirected", diag = FALSE)


ggraph(K50, "stress") +
  geom_edge_link0(edge_linewidth = 0.6, edge_color = "grey66") +
  geom_node_point(shape = 21, fill = "grey66", size = 2, show.legend = FALSE) +
  theme_void() +
  coord_fixed()
```

```{r}
clu_louvain <- cluster_louvain(K50)
table(membership(clu_louvain))
```

In such cases, can use `cluster_leiden` instead. When used with the CPM (Constant Potts Model) objective function, it avoids the resolution limit entirely.

```{r}
clu_leiden <- cluster_leiden(
  K50, objective_function = "CPM", resolution = 0.5
)
table(membership(clu_leiden))
```

Choosing the right resolution can be tricky. A common approach is to calculate

```{r}
(r <- quantile(degree(karate))[2] / (vcount(karate) - 1))
```

```{r}
leidenc <- cluster_leiden(
  karate,
  objective_function = "CPM",
  resolution = r
)
```

```{r}
#| label: fig-karate-leiden
#| fig-cap: "Karate club network colored by community membership (Leiden algorithm with CPM objective function)."

V(karate)$clu <- membership(leidenc)
ggraph(karate, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(
    shape = 21,
    size = 5,
    aes(fill = as.factor(clu)),
    show.legend = FALSE
  ) +
  theme_void() +
  coord_equal()
```

# Blockmodeling

Focuses on patterns of connections in addition to density. Two nodes belong to the same block if they relate to other blocks in similar ways, whether or not they are connected to each other.  This makes blockmodeling particularly useful for role analysis in sociology and organizational context.

Blockmodeling is computationally intensive, and suited for small networks.

We illustrate blockmodeling on a random network with 3 dense blocks of size 20.

```{r}
#| label: fig-blockmodel-network
#| fig-cap: "A random network with three dense blocks of 20 nodes each."

n1 <- 3
n2 <- 20

set.seed(1108)
g <- sample_islands(n1, n2, 0.75, 5)
g <- simplify(g)
V(g)$grp <- rep(LETTERS[1:n1], each = n2)
ggraph(g, "stress") +
  geom_edge_link0(edge_linewidth = 0.2, edge_color = "grey66") +
  geom_node_point(shape = 21, size = 5, fill = "grey66", show.legend = FALSE) +
  theme_void()
```

Must define expected block structure with an adjacency matrix specifying expectations for each pair of blocks. "com" is complete with all possible ties, "nul" means no ties expected.  In the example below, we place "com" on the diagonal and "nul" on the off-diagonal, which corresponds to the familiar clustering scenario where nodes are densely connected within groups and sparsely connected between them. The function then tries multiple random starting partitions (rep) and optimizes each one, returning the best assignment of nodes to blocks.

```{r}
A <- as_adjacency_matrix(g)
(blk <- matrix(
  c("com", "nul", "nul", "nul", "com", "nul", "nul", "nul", "com"),
  nrow = 3
))
```

```{r}
res <- optRandomParC(
  M = A, k = 3,
  approaches = "bin", #binary blockmodel
  blocks = blk, 
  rep = 5, # random starting partitions
  mingr = 20, maxgr = 20 #min and max block size
)
```

```{r}
clu(res)
```

```{r}
#| label: fig-blockmodel-result
#| fig-cap: "Blockmodeling result for the random network."

V(g)$clu <- clu(res)
ggraph(g, "stress") +
  geom_edge_link0(edge_linewidth = 0.2, edge_color = "grey66") +
  geom_node_point(
    shape = 21,
    size = 5,
    aes(fill = as.factor(clu)),
    show.legend = FALSE
  ) +
  theme_void()
```

# Core-Perifery

Unlike community detection, core-perifery posits that a network is organized into a densely connected core and a sparsely-connected periphery. Core nodes are highly connected to each other and periphery nodes, periphery nodes tend to be connected to core nodes but not each other.

```{r}
#| label: fig-compare-clu-cp
#| fig-cap: "Comparison of community structure and core-periphery structure by patterns in the adjacency matrix. Left: community structure with dense blocks along the diagonal. Right: core-periphery structure with a dense block in the top-left corner and sparse connections elsewhere."

set.seed(1108)
g_clu <- sample_islands(3, 8, 0.8, 2)
V(g_clu)$clu <- rep(1:3, each = 8)
g_clu <- permute(g_clu, order(V(g_clu)$clu))
V(g_clu)$clu <- rep(1:3, each = 8)
E(g_clu)$type <- apply(
  matrix(V(g_clu)$clu[as_edgelist(g_clu, names = FALSE)], ncol = 2),
  1,
  function(x) if (x[1] == x[2]) paste("community", x[1]) else "between"
)

p1 <- ggraph(g_clu, "matrix") +
  geom_edge_point(mirror = TRUE, aes(edge_color = type)) +
  coord_fixed() +
  scale_y_reverse() +
  theme_graph() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  labs(title = "Community structure")

set.seed(1108)
g_cp <- netUtils::split_graph(n = 24, p = 0.3, core = 0.3)
V(g_cp)$type <- c("periphery", "core")[V(g_cp)$core + 1]
E(g_cp)$type <- apply(
  matrix(V(g_cp)$core[as_edgelist(g_cp, names = FALSE)], ncol = 2),
  1,
  function(x) if (all(x == 1)) "core - core" else "core - periphery"
)

p2 <- ggraph(g_cp, "matrix") +
  geom_edge_point(mirror = TRUE, aes(edge_color = type)) +
  coord_fixed() +
  scale_y_reverse() +
  theme_graph() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  labs(title = "Core-periphery structure")

p1 + p2
```

Core-periphery structures can appear in many different contexts. In international trade networks, a small number of wealthy, industrialised countries form a densely interconnected core, while developing nations at the periphery trade primarily with core countries but rarely with each other. In scientific collaboration networks, a core of highly productive researchers co-author with many others and with each other, while peripheral researchers have fewer and more localised collaborations. 

The idea of the model is that nodes either belong to the core, or the periphery. The membership of nodes is derived by optimizing the correlation between the observed adjacency matrix and an idealized adjacency matrix.

```{r}
#| label: fig-core-graph
#| fig-cap: "A graph with a perfect core-periphery structure. Left: network layout with core and periphery nodes. Right: adjacency matrix showing the block pattern."

cp <- core_periphery(core_graph)
V(core_graph)$type <- c("periphery", "core")[cp$vec + 1]
p1 <- ggraph(core_graph, "stress") +
  geom_edge_link0(edge_color = "grey66", edge_linewidth = 0.1) +
  geom_node_point(shape = 21, size = 3, aes(fill = type)) +
  theme_graph() +
  theme(legend.position = "bottom", legend.title = element_blank())

E(core_graph)$col <- apply(
  matrix(as.logical(cp$vec[as_edgelist(core_graph, names = FALSE)]), ncol = 2),
  1,
  all
)
E(core_graph)$type <- c("core - periphery", "core - core")[
  E(core_graph)$col + 1
]
p2 <- ggraph(core_graph, "matrix") +
  geom_edge_point(mirror = TRUE, aes(edge_color = type), edge_size = 0.2) +
  coord_fixed() +
  scale_y_reverse() +
  guides(edge_color = guide_legend(override.aes = list(edge_size = 3))) +
  theme_graph() +
  theme(legend.position = "bottom", legend.title = element_blank())

p1 + p2
```

```{r}
(cp <- core_periphery(core_graph))
```

`vec` is one if core, zero if peripheral. `corr` is the correlation to the ideal adjacency matrix derived from `vec` membership, which in this case is perfect.

Different optimization techniques can be used. To illustrate this on a real-world network, we use the US domestic flights network from the `networkdata` package. This network connects 276 airports based on scheduled flights and is a natural candidate for core-periphery analysis: major hub airports (Atlanta, Denver, Dallas-Fort Worth, etc.) serve as a densely interconnected core, while smaller regional airports on the periphery connect primarily to these hubs.

```{r}
cp_dc <- core_periphery(us_flights, method = "rk1_dc")
cp_dc$corr; sum(cp_dc$vec)
```

45 core airports identified, with a correlation of 0.81, strongly suggesting a core-periphery structure in the airline system. The `GA` method is usually better, but much slower. In this case, rank-one approximations are almost as good.

```{r}
cp_ec <- core_periphery(us_flights, method = "rk1_ec")
cp_ga <- core_periphery(us_flights, method = "GA", iter = 2500)
c(rk1_dc = cp_dc$corr, rk1_ec = cp_ec$corr, GA = cp_ga$corr)
```









