---
title: "Two-Mode Networks"
format: gfm
---

[Source](https://schochastics.github.io/R4SNA/descriptive/two-mode-networks.html)

```{r}
options(paged.print = FALSE)
```

```{r}
#| message: false

libraries <- list(
  "igraph", "networkdata", "tnet",
  "backbone", "ggraph", "tidyverse"
)
invisible(lapply(libraries, library, character.only = TRUE))
```


Two-mode network data represent situations like

- Affiliation networks (membership in institutions/clubs)
- Voting/sponsorship networks (politicians and bills)
- Citation networks (authors and papers)
- Co-authorship networks (also authors and papers)

The can be analyzed directly or projected onto a one-mode network, the latter approach being more popular.

# Data Structure

Stored as an _incidence matrix_ or _biadjacency matrix_, with rows for one type of node and columns for the other type. Entries indicated ties existing across the two sets. Two-mode networks are especially useful for analyzing how connections emerge indirectly, for example, when two individuals are linked because they attend the same event or belong to the same organization.

`igraph` recognizes a two-mode network if it contains a logical attribute named `type`. One mode is `FALSE` and the other is `TRUE`.

To illustrate methods tailored for two-mode networks, we will use the well-known Southern Women dataset (Davis et al. 2009). This classic dataset records the attendance patterns of 18 women at 14 social events in the American South, and has become a foundational example in network analysis. 

```{r}
southern_women
```

```{r}
(A <- as_biadjacency_matrix(southern_women))
```

```{r}
#| label: fig-incmat
#| message: false
#| fig-cap: "Incidence matrix representation of the Southern Women two-mode network. Rows correspond to the 18 women and columns to the 14 social events; black cells indicate attendance at an event, and white cells indicate non-attendance."

# Convert to tidy format
df <- as.data.frame(A) %>% 
  rownames_to_column("woman") %>% 
  pivot_longer(-woman, names_to = "event", values_to = "attend") %>% 
  mutate(
    attend = as.integer(attend > 0),
    # Reverse row order so first woman appears at top
    woman = factor(woman, levels = rev(unique(woman))),
    event = factor(event, levels = unique(event))
  )

# Plot incidence matrix (black = tie, white = no tie)
ggplot(df, aes(x = event, y = woman, fill = factor(attend))) +
  geom_tile(color = "grey70", linewidth = 0.3) +
  scale_fill_manual(values = c("0" = "white", "1" = "black"), guide = "none") +
  coord_fixed() +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

```{r}
#| label: fig-southern-women-plot
#| fig-height: 7
#| fig-width: 9

ggraph(southern_women, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(
    aes(fill = type, shape = type),
    size = 8,
    show.legend = FALSE
  ) +
  geom_node_text(aes(label = name)) +
  scale_shape_manual(values = c(21, 22)) +
  theme_void()
```

To see how many vertices belong to each mode:

```{r}
table(V(southern_women)$type)
```

# Direct Approach

Since closed triplets of nodes do not exist, the definition of transitivity must be adapted.

```{r}
transitivity(southern_women); transitivity(southern_women, type = "local")
```

`tnet` defines a two-mode clustering coefficient based on cycles of length six, eg. woman $\rightarrow$ event $\rightarrow$ woman $\rightarrow$ event $\rightarrow$ woman $\rightarrow$ event $\rightarrow$ back to the first woman.

```{r}
el_women <- as_edgelist(southern_women, names = F)
```

Global clustering (first mode)

```{r}
clustering_tm(el_women)
```

Local clustering (first mode).

```{r}
clustering_local_tm(el_women)
```

Local clustering (second mode).

```{r}
clustering_local_tm(el_women[, 2:1])
```

`NaN` values indicated the node does not participate in any valid six-cycle. This approach is computationally intensive.

# Projection Approach

Nodes from one mode are connected if they share ties to the same nodes in the other mode. They can be weighted or binary.

## Weighted Projection

Projection onto the first mode is $AA^T$, projecting onto the second mode is $A^TA$. The result is a weighted adjacency matrix, with entries indicating the number of shared affiliations.

```{r}
(B <- A %*% t(A))
```

Alternatively, with `igraph`

```{r}
(projs <- bipartite_projection(southern_women))
```

```{r}
#| label: fig-plot_weighted_proj
#| fig-height: 8
#| fig-width: 10
#| fig-cap: "Weighted projection of the Southern Women network onto the women."

proj <- graph_from_adjacency_matrix(
  B,
  weighted = TRUE,
  diag = FALSE,
  mode = "undirected"
)

ggraph(proj, "stress") +
  geom_edge_link0(
    aes(edge_linewidth = weight),
    edge_color = "grey66",
    show.legend = FALSE
  ) +
  geom_node_point(shape = 21, fill = "tomato", size = 8, show.legend = FALSE) +
  geom_node_text(aes(label = name)) +
  scale_edge_width(range = c(1, 4)) +
  theme_graph() +
  coord_cartesian(clip = "off")
```

This is commonly binarized for analysis.

## Simple Binary Projections

A simple way is to define a global threshold for turning weights into binary. This can be the mean edge weight, sometimes adjusted upwards by 1-2 standard deviations. In many projected networks, the distribution of edge weights is highly skewed, with most pairs sharing only one or two affiliations. 

The appropriate threshold depends on the research question: lower thresholds preserve more information but may produce very dense graphs, while higher thresholds emphasize strong ties but risk discarding meaningful structure. 

```{r}
women_proj <- projs$proj1
threshold <- mean(E(projs$proj1)$weight)
women_bin <- delete_edges(
  women_proj, which(E(women_proj)$weight <= threshold)
)
(women_bin <- delete_edge_attr(women_bin, "weight"))
```

Connections now only show stronger-than-average shared-event ties.

```{r}
#| label: fig-southern-women-bin-plot
#| fig-cap: "Binary projection of the Southern Women network onto the women using a global threshold based on the mean edge weight."
#| fig-height: 11
#| fig-width: 11

ggraph(women_bin, "stress", bbox = 5) +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(shape = 21, fill = "tomato", size = 8) +
  geom_node_text(aes(label = name)) +
  theme_graph() +
  coord_cartesian(clip = "off")
```

## Model-based Binary Projections

Statistical models which determine if an edge weight differs enough from the expected value of an underlying null model to keep the edge in the binary projection.

The idea is:

1. Create the weighted projection of interest, eg. `B <- A%*%t(A)`
2. Generate random two-mode networks according to a given model
3. Compare if the values `B[i,j]` differ significantly from the distribution of values in the random projections.

The difference is in the construction of the random networks:

- _Fixed Degree Sequence Model_ `fdsm`: Create random two-mode networks with the same row and column sums as `A`.
- _Fixed Column Model_ `fixedcol`: Create random two-mode networks with the same column sums as `A`.
- _Fixed Row Model_ `fixedrow`: Create random two-mode networks with the same row sums as A.
- _Fixed Fill Model_ `fixedfill`: Create random two-mode networks with the same number of ones as A.
- _Stochastic Degree Sequence Model_ `sdsm`: Create random two-mode networks with approximately the same row and column sums as A.

In general you can follow these rough questions:

1. Use the model that fits your empirical setting or a known link formation process. If that link formation process dictates that row sums are fixed but column sums not, then choose fixedrow.
2. Use fdsm if your network is small enough. Sampling from the FDSM is quite expensive.
3. Use the sdsm for large networks.

To illustrate model fitting in a substantive setting, we use a bill cosponsorship network from the U.S. Senate in 2015 (available in the networkdata package). The data form a two-mode network in which one set of nodes represents senators and the other represents bills; a tie between a senator and a bill indicates sponsorship or cosponsorship. We are interested in examining the binary one-mode projection onto senators, i.e., a network in which two senators are connected if they sponsored at least one bill together.

```{r}
cosponsor
```

Because the network is large, use SDSM. The projection is onto the nodes with `type = FALSE`. For `type = TRUE`, the `type` attribute must be inverted.

```{r}
(senators <- backbone_from_projection(
  cosponsor,
  model = "sdsm",
  alpha = 0.05,
  signed = FALSE
))
```

If `signed = FALSE`, ties are retained only if the co-sponsorship strength is greater tahn expected from a null model at the significance level `alpha`. If `signed = TRUE`, negative results for less than expected strength of co-sponsorship, with positive numbers showing greater than expected.

```{r}
#| label: fig-cosponsor-plot
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Backbone projection of the U.S. Senate cosponsorship network, with nodes colored by party affiliation. Edges represent statistically significant co-sponsorship ties based on the SDSM null model."

V(senators)$party <- V(cosponsor)$party[match(
  V(senators)$name,
  V(cosponsor)$name
)]
senators <- largest_component(senators)
party_cols <-
  c(D = "#104E8B", R = "#CD3333", I = "#EEB422")
ggraph(senators, "stress") +
  geom_edge_link0(edge_linewidth = 0.1, edge_color = "grey66") +
  geom_node_point(shape = 21, aes(fill = party), size = 4) +
  scale_fill_manual(values = party_cols) +
  theme_graph() +
  theme(legend.position = "bottom")
```

This reveals strong polarization, with a small number of edges spanning the gap, and a few senators positioned as "brokers".







