---
title: "Ego Networks"
format: gfm
---

[Source](https://schochastics.github.io/R4SNA/descriptive/ego-networks.html)

```{r}
options(paged.print = FALSE)
```

```{r}
#| message: false

libraries <- list(
  "igraph", "networkdata", "egor",
  "patchwork", "ggraph", "tidyverse",
  "knitr"
)
invisible(lapply(libraries, library, character.only = TRUE))
```

# Data Structure

Ego network analysis focuses on a single individual (ego), people directly connected to them (alters), and the relation ship among the alters.

Data consists of multiple distinct networks, each centered on a different ego and typically composed of different actors.

The `egor` object stores three levels:

- _ego-level data_ - attributes of respondents
- _alter-level data_ - attributes of individuals named by respondents
- _alter-alter ties_ - relationships among alters

# Create the UCNets Ego Network Dataset

In this chapter, we draw on ego network data from the UC Berkeley Social Networks Study (UCNets) (Fischer 2020), a longitudinal study of personal networks in the San Francisco Bay Area conducted between 2015 and 2018. Specifically, we use data from the first wave and select only a subset of variables relevant for the examples presented in Chapter 7. Several of these variables are recoded to create analytically useful ego-, alter-, and alter–alter-level measures.

To construct the ego-centered network dataset used in Chapter 7, we combine information from two wave-1 files: one containing respondent-level information and one containing alter- and alter–alter-level information. These are then organized into the three components required by the egor package: an ego table, an alter table, and an alter–alter tie table.

## Load Wave 1 Data

```{r}
alt_obj <- load("data/ICPSR_36975/DS0001/36975-0001-Data.rda")
altw1 <- get(alt_obj)

ego_obj <- load("data/ICPSR_36975/DS0005/36975-0005-Data.rda")
egow1 <- get(ego_obj)

# Alter-alter pair variables in the alter file
pair_vars <- c(
  "N1_N2", "N1_N3", "N1_N4", "N1_N5",
  "N2_N3", "N2_N4", "N2_N5",
  "N3_N4", "N3_N5",
  "N4_N5"
)
```

## Ego-level data

```{r}
egos <- egow1 %>% 
  select(
    PRIM_KEY, GENDER, YEAR_PRELOAD, AGE_EX, A2A, A13A, A19A, B3B,
    G1, H3, H4A, H5A, H6A, J1A, J1B, K1, RACECATS1, K9A, K15, K22A
  ) %>%
  mutate(
    .egoID = as.character(PRIM_KEY),

    # 0 = male, 1 = female
    gender = ifelse(as.numeric(GENDER) == 2, 1, 0),

    year_birth = YEAR_PRELOAD,

    # exact age in years
    age = as.numeric(AGE_EX),

    # 1 = married, 0 = otherwise
    married = ifelse(as.numeric(A2A) == 1, 1, 0),

    # 1 = yes, 0 = no
    pets = ifelse(as.numeric(A13A) == 1, 1, 0),

    # 1 = own, 0 = rent/other
    home_owner = ifelse(as.numeric(A19A) == 1, 1, 0),

    # 1 = active in informal groups, 0 = not active
    informal_groups = ifelse(as.numeric(B3B) == 1, 1, 0),

    # ordered self-rated health
    health = ordered(
      as.numeric(G1),
      levels = c(1, 2, 3, 4, 5),
      labels = c("excellent", "very_good", "good", "fair", "poor")
    ),

    # 1 = advised to lose weight, 0 = otherwise
    lose_weight_advice = ifelse(as.numeric(H3) == 1, 1, 0),

    # 1 = smoker, 0 = non-smoker
    smoker = ifelse(as.numeric(H4A) == 1, 1, 0),

    # days per week drinking alcohol
    drinking_days = as.numeric(H5A),

    # 1 = used marijuana, 0 = otherwise
    marijuana_use = ifelse(as.numeric(H6A) == 1, 1, 0),

    # extraversion = reverse-coded reserved + sociable
    extraversion = rowMeans(
      cbind(
        6 - as.numeric(J1A),
        as.numeric(J1B)
      ),
      na.rm = TRUE
    ),

    # ordered education
    education = ordered(
      as.numeric(K1),
      levels = 1:10,
      labels = c(
        "less_than_9th",
        "9th_to_12th_no_grad",
        "high_school",
        "ged",
        "some_college",
        "associate",
        "bachelor",
        "master",
        "professional_degree",
        "other"
      )
    ),

    # race
    race = factor(
      as.numeric(RACECATS1),
      levels = c(1, 2, 3, 4, 6, 7, 9),
      labels = c(
        "white",
        "black",
        "american_indian",
        "asian",
        "unknown",
        "pacific_islander",
        "mixed"
      )
    ),

    # country of birth collapsed to continent
    birth_continent = factor(
      as.numeric(K9A),
      levels = c(1, 2, 4, 6, 11, 3, 7, 8, 9, 10, 12, 14, 5, 13, 15),
      labels = c(
        "north_america", "north_america", "north_america",
        "north_america", "north_america",
        "asia", "asia", "asia", "asia", "asia", "asia", "asia",
        "europe", "europe",
        "other"
      )
    ),

    # religion
    religion = factor(
      as.numeric(K15),
      levels = c(1, 2, 3, 4, 5, 6, 7),
      labels = c(
        "protestant",
        "catholic",
        "jewish",
        "muslim",
        "buddhist",
        "other_religion",
        "no_religion"
      )
    ),

    # ordered income
    income = ordered(
      as.numeric(K22A),
      levels = 1:13,
      labels = c(
        "under_15k",
        "15k_25k",
        "25k_35k",
        "35k_45k",
        "45k_60k",
        "60k_75k",
        "75k_100k",
        "100k_125k",
        "125k_150k",
        "150k_200k",
        "200k_300k",
        "300k_500k",
        "500k_plus"
      )
    ), 
    .keep = "none"
  ) %>% 
  distinct()
```



## Alter-level data

```{r}
alters <- altw1 |>
  filter(!is.na(PRIM_KEY), !is.na(NAME_NMBR2)) |>
  transmute(
    .egoID = as.character(PRIM_KEY),
    .altID = as.integer(NAME_NMBR2),

    # match ego coding: 0 = male, 1 = female
    gender = ifelse(as.numeric(N_GENDER) == 2, 1, 0),

    # family indicator
    family = ifelse(
      C1A_1 == "(1) yes" |
      C1A_3 == "(1) yes" |
      C1A_4 == "(1) yes" |
      C1A_5 == "(1) yes" |
      C1A_6 == "(1) yes",
      1, 0
    ),

    # race homophily indicator
    same_race = ifelse(C2G_NSAMERACE == "(1) yes", 1, 0),

    # alter roughly same age as ego (+/- 6 years)
    same_age = case_when(
      as.numeric(C1C_NSAMEAGE) == 1 ~ 1,
      as.numeric(C1C_NSAMEAGE) == 0 ~ 0,
      TRUE ~ NA_real_
    )
  ) |>
  distinct()
```

## Alter-alter ties

```{r}
aaties <- altw1 |>
  select(PRIM_KEY, all_of(pair_vars)) |>
  distinct() |>
  pivot_longer(
    cols = all_of(pair_vars),
    names_to = "pair",
    values_to = "relation"
  ) |>
  filter(!is.na(relation)) |>
  mutate(
    relation = as.character(relation),

    # tie if alters know each other very well or know a little
    weight = ifelse(grepl("^\\((1|2)\\)", relation), 1, 0),

    .egoID = as.character(PRIM_KEY),
    .srcID = as.integer(sub("^N([1-5])_N([1-5])$", "\\1", pair)),
    .tgtID = as.integer(sub("^N([1-5])_N([1-5])$", "\\2", pair))
  ) |>
  filter(weight == 1) |>
  select(.egoID, .srcID, .tgtID, weight)

valid_alters <- alters |>
  select(.egoID, .altID)

aaties <- aaties |>
  inner_join(valid_alters, by = c(".egoID", ".srcID" = ".altID")) |>
  inner_join(valid_alters, by = c(".egoID", ".tgtID" = ".altID"))
```

## Build ego object

```{r}
ego_net_w1 <- egor(
  egos   = egos,
  alters = alters,
  aaties = aaties,
  ID.vars = list(
    ego    = ".egoID",
    alter  = ".altID",
    source = ".srcID",
    target = ".tgtID"
  )
)
```

```{r}
head(ego_net_w1$ego)
head(ego_net_w1$alter)
head(ego_net_w1$aatie)
```

Before proceeding with the analysis, we remove respondents for whom the variable numgiven is missing in the ego-level data. This variable records how many alters each respondent named and is required for several of the descriptive measures used later in the chapter. Because numgiven records how many alters a respondent named, a missing value at the ego level implies that no alter information is recorded for that respondent. Consequently, it is sufficient to remove missing values only from the ego table, since the alter and alter–alter tie tables contain entries only for egos who actually reported network members.

```{r}
dim(ego_net_w1$ego); dim(ego_net_w1$alter); dim(ego_net_w1$aatie)
```

The ego table now contains 1159 respondents. Across these respondents, a total of 12209 alters were named, each linked to a specific ego via .egoID and uniquely identified within each ego network by .altID. The alter–alter tie table contains 6696 rows, with each row representing a relationship between two alters within the same ego network. These ties are identified by the .egoID of the corresponding ego and the .srcID and .tgtID variables that indicate the two alters involved.

# Plotting Ego Nets

Convert to `igraph` object and view first 3 ego networks. Each list item contains one ego's personal network. Alter-level attributes are automatically transferred as vertex attributes, and alter-alter tie information as edge attributes.

```{r}
g <- as_igraph(ego_net_w1)
g[1:3]
```

```{r}
#| label: fig-egos
#| message: false
#| fig-height: 9
#| fig-width: 9
#| fig-cap: "The first nine ego networks derived from the egor object based on the example dataset. Each panel represents the alter–alter structure of a single ego's personal network, where nodes correspond to alters and edges represent ties among them. The ego node itself is not shown, as it is by definition connected to all alters. Differences across panels illustrate variation in density, clustering, and overall structural configuration across ego networks."

plots <- lapply(g[1:9], function(g) {
  # rebuild a clean graph using only structure + names
  el <- as_edgelist(g, names = TRUE)
  vn <- V(g)$name

  if (length(el) == 0 || nrow(el) == 0) {
    g_clean <- make_empty_graph(n = length(vn), directed = FALSE)
    V(g_clean)$name <- vn
  } else {
    g_clean <- graph_from_edgelist(el, directed = FALSE)

    # add back isolated vertices, if any
    missing_v <- setdiff(vn, V(g_clean)$name)
    if (length(missing_v) > 0) {
      g_clean <- add_vertices(g_clean, nv = length(missing_v), name = missing_v)
    }
  }

  ggraph(g_clean, layout = "kk") +
    geom_edge_link(alpha = 0.4) +
    geom_node_point(size = 6, color = "steelblue") +
    geom_node_text(aes(label = name), size = 3, color = "white") +
    theme_graph()
})

wrap_plots(plots, ncol = 3)
```

# Descriptive Statistics

_Density_ captures the proportion of actual ties to all possible ties. It reflects the cohesion within a personal network. High values indicate tightly knit groups. Such cohesive structures can facilitate trust, norm enforcement, and social support, as information circulates quickly and reputational mechanisms are strong.

OTOH, dense networks tend to generate redundant information because alters are connected to many of the same people and share overlapping perspectives. Individuals benefit when they bridge gaps, so-called structural holes, between otherwise disconnected groups. In low-density networks, ego may occupy a brokerage position linking clusters that are not directly connected to each other, thereby gaining access to diverse, non-redundant information and potential strategic advantages (such as faster promotion within organizations). Density therefore captures an important theoretical trade-off between cohesion and brokerage in ego network structure.

```{r}
dens <- ego_density(ego_net_w1)
head(dens)
```

Note ego networks with 0 or 1 alters will have density of `NaN`.

```{r}
sum(is.nan(dens$density))
```

```{r}
#| label: fig-ego-dens
#| fig-cap: "Distribution of densities across ego networks. Values closer to 1 indicate networks in which most alters are connected to one another, while values closer to 0 indicate networks with many structural holes."
#| warning: false

ggplot(dens, aes(x = density)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white") +
  labs(
    x = "Ego network density",
    y = "Number of ego networks"
  ) +
  theme_minimal()
```

## Network composition based on alter attributes

The _composition of alters_ focuses on who is present in the ego's personal network by aggregating alter attributes within each ego network.

`comp_ply` applies a function to an alter attribute within each ego network, returning one value per ego.

### Proportion of Family Members

This captures the extent to which a respondent’s personal network is centered around kin relationships rather than non-kin ties such as friends, coworkers, or acquaintances. Networks with a high proportion of family members may indicate stronger reliance on kin-based support systems, whereas networks with fewer family members may reflect more diverse social connections outside the household or extended family.

```{r}
prop_family <- comp_ply(ego_net_w1, "family", mean)
head(prop_family)
```

### Gender composition

Values close to 0 indicate networks composed mostly of men, values close to 1 indicate networks composed mostly of women, and values around 0.5 suggest a more gender-balanced network.

```{r}
prop_female <- comp_ply(ego_net_w1, "gender", mean)
head(prop_female)
```

```{r r4s-ego-1}
ggplot(prop_female, aes(x = result))  +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white") +
  labs(
    x = "Ego network density",
    y = "Number of ego networks"
  ) +
  theme_minimal()
```

### Racial Homiphily

```{r}
prop_same_race <- comp_ply(ego_net_w1, "same_race", mean)
head(prop_same_race)
```

```{r}
comp_summary <- data.frame(
  egoID = prop_family$.egoID,
  prop_family = prop_family$result,
  prop_female = prop_female$result,
  prop_same_race = prop_same_race$result
)

row.names(comp_summary) <- NULL

kable(
  head(comp_summary, 10),
  digits = 2,
  caption = "Summary of compositional measures for ego networks."
)
```

## Diversity of alters within ego networks

Instead of the presence of certain alter characterstics, diversity captures how heterogenous the alters are with respect to a given attribute. Shannon entropy

$$
H = - \sum_i p_i \log(p_i)
$$

where $p_i$ represents the proportion of alters belonging to category $i$. Entropy increases when alters are more evenly distributed across categories and decreases when most alters belong to the same category. Networks with entropy close to zero therefore contain alters who are largely similar to one another, while higher values indicate more heterogeneous networks.

```{r}
gender_entropy <- alts_diversity_entropy(
  ego_net_w1,
  alt.attr = "gender",
  base = exp(1)
)
head(gender_entropy)
```

Networks composed of mostly the same gender have lower values.

We can combine the diversity measure with the compositional measures above to better understand how ego networks vary across respondents.

```{r r4s-ego-2}
comp_summary$gender_entropy <- gender_entropy$entropy

comp_long <- comp_summary %>% 
  pivot_longer(
    cols = -egoID,
    names_to = "measure",
    values_to = "value"
  )

comp_long %>% 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~measure, ncol = 2, scales = "free") +
  labs(
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal()  
```

The proportion of family members varies widely across respondents, indicating substantial heterogeneity in the extent to which personal networks are centered around kin. The gender composition measure is broadly distributed around intermediate values, suggesting that many respondents’ networks contain a mixture of male and female alters rather than being strongly gender-segregated. The gender diversity measure is concentrated toward higher values, indicating that many ego networks contain a relatively balanced mix of male and female alters. Finally, the distribution of racial homophily is highly skewed toward 1, suggesting that most respondents name alters who share their racial background.

# Predicting individual outcomes using ego network measures

Since ego network data are typically collected through individual surveys, the resulting observations can be treated as independent, allowing for traditional regression models.

## Predicting self-reported health - Ordinal Regression

> Are respondents with larger, more diverse, or more kin-centered personal networks more likely to report better health?

### Outcome Variable

The outcome is the ego-level variable `health`. The variable is an ordered factor with five categories.

The model will use an ordinal regression model.

### Predictors

Ego-level `age`, `gender` and `education` and ego-network measures `ego_size`, `prop_family` and `gender_entropy`. `prop_same_race` has too many missing observations.

Together, this model combines network size, network composition, homophily, and diversity, along with standard demographic controls.

### Preparing Analysis Data

```{r}
net_size <- ego_net_w1$alter %>% 
  count(.egoID, name = "ego_size")

prop_family_df <- comp_ply(ego_net_w1, "family", mean) %>% 
  rename(prop_family = result)
prop_same_race_df <- comp_ply(ego_net_w1, "same_race", mean) %>% 
  rename(prop_same_race = result)
gender_entropy_df <- alts_diversity_entropy(
  ego_net_w1, alt.attr = "gender", base = exp(1)
) %>% 
  rename(gender_entropy = entropy)

analysis_dat <- egos %>% 
  left_join(net_size, by = ".egoID") %>% 
  left_join(prop_family_df, by = ".egoID") %>% 
  left_join(prop_same_race_df, by = ".egoID") %>% 
  left_join(gender_entropy_df, by = ".egoID")
head(analysis_dat)  
```

```{r}
analysis_dat$health_ord <- ordered(
  analysis_dat$health,
  levels = c("poor", "fair", "good", "very_good", "excellent")
)

table(analysis_dat$health_ord, useNA = "ifany")
```

### Estimating the model

```{r}
library(MASS)

analysis_dat$education <- factor(analysis_dat$education)

health_model <- polr(
  health_ord ~ ego_size + prop_family + gender_entropy +
    age + gender + education,
  data = analysis_dat, Hess = TRUE, na.action = na.omit
)

summary(health_model)
```

### Interpretation

In ordinal logistic regression estimated with polr(), statistical significance can be assessed by examining the ratio of the coefficient to its standard error (the value). As a rule of thumb, absolute values larger than about 1.96 correspond to significance at the 5% level under the normal approximation. Applying this criterion, the effect of network size is statistically significant, suggesting that respondents with larger personal networks are more likely to report better health categories. Education is also positively correlated and statistically significant. None of the other measures are.

### Predicted Probabilities

Although the coefficients from an ordinal logistic regression indicate how predictors shift respondents along the underlying latent scale of the outcome, they can be difficult to interpret substantively. A more intuitive way to understand the model is to examine predicted probabilities for each category of the outcome variable.

For example, we can examine how the predicted probability of reporting different health levels changes as ego network size increases, while holding the other variables constant.

```{r}
pred_probs <- predict(health_model, type = "probs")
head(pred_probs)
```

To visualize the relationship between network size and health, we can generate predicted probabilities across a range of ego network sizes while holding the other predictors at typical values.

```{r r4s-ego-3}
size_seq <- seq(
  min(analysis_dat$ego_size, na.rm = T),
  max(analysis_dat$ego_size, na.rm = T),
  length.out = 50
)

pred_dat <- tibble(
  ego_size = size_seq,
  prop_family = mean(analysis_dat$prop_family, na.rm = T),
  gender_entropy = mean(analysis_dat$gender_entropy, na.rm = T),
  age = mean(analysis_dat$age, na.rm = T),
  gender = 0,
  education = levels(analysis_dat$education)[1]
)

pred_dat <- cbind(
  pred_dat,
  predict(health_model, newdata = pred_dat, type = "probs")
)

pred_long <- pred_dat %>% 
  pivot_longer(
    cols = c(poor, fair, good, very_good, excellent),
    names_to = "health_category",
    values_to = "probability"
  )

pred_long %>% 
  ggplot(aes(x = ego_size, y = probability, color = health_category)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Ego Network Size",
    y = "Predicted Probability",
    color = "Health Category"
  ) +
  theme_minimal()  
```

Overall, the results show a clear pattern: as the number of alters in a respondent’s personal network increases, the probability of reporting better health categories increases, while the probability of reporting poorer health categories declines. This pattern indicates that larger ego networks are associated with a greater likelihood of reporting better health outcomes.

Substantively, this finding aligns with the idea that individuals embedded in larger personal networks may benefit from greater social support, access to information, and other resources that can contribute to improved health and well-being.

## Predicting Smoking Behavior - Logistic Regression

We now illustrate a second application by examining whether features of respondents’ ego networks are associated with a binary health behavior outcome: cigarette smoking.

Smoking behavior is a natural outcome to consider in a network context. Social networks often shape health behaviors through mechanisms such as social norms, peer influence, and access to support. Individuals embedded in larger or more diverse networks may experience different pressures or resources that affect their likelihood of smoking. For example, networks with strong family presence may discourage smoking, while more diverse networks may expose individuals to different behavioral norms.

> Are respondents with larger, more diverse, or more kin-centered personal networks more or less likely to smoke cigarettes?

### Outcome variable

`smoker`, coded as 1 = smoker. The model will be a logistical regression model.

### Predictors

As before

### Estimating the Model

```{r}
smoke_model <- glm(
  smoker ~ ego_size + prop_family + gender_entropy +
    age + gender + education,
  data = analysis_dat,
  family = binomial, na.action = na.omit
)
summary(smoke_model)
```

### Interpretation

Among the network predictors, ego network size (ego_size) shows a negative association with smoking. The coefficient is −0.061, corresponding to an odds ratio of approximately 0.94. This suggests that each additional alter in a respondent’s ego network is associated with roughly a 6% decrease in the odds of smoking, although the effect is only marginally significant. 

Among the demographic controls, gender shows a marginally significant association with smoking. The negative coefficient indicates that respondents coded as female have lower odds of smoking than males. 

Finally, the results indicate that education is associated with smoking behavior, as reflected in the statistically significant polynomial contrasts for the education variable. These contrasts capture systematic variation in smoking across levels of educational attainment. Substantively, this pattern is consistent with well-established findings that smoking prevalence tends to be lower among individuals with higher levels of education.

For ease of interpretation, it is often useful to transform the coefficients into odds ratios, which indicate the multiplicative change in the odds of smoking associated with a one-unit increase in the predictor. Odds ratios greater than 1 indicate that the predictor is associated with higher odds of smoking, while values below 1 indicate lower odds of smoking.

```{r}
exp(coef(smoke_model))
```

### Predicted Probabilities

How the probability of smoking changes as ego network size increases, while holding the other predictors constant.


```{r r4s-ego-4}
size_seq <- seq(
  min(analysis_dat$ego_size, na.rm = TRUE),
  max(analysis_dat$ego_size, na.rm = TRUE),
  length.out = 50
)

pred_dat <- data.frame(
  ego_size = size_seq,
  prop_family = mean(analysis_dat$prop_family, na.rm = TRUE),
  gender_entropy = mean(analysis_dat$gender_entropy, na.rm = TRUE),
  age = mean(analysis_dat$age, na.rm = TRUE),
  gender = 0,
  education = levels(analysis_dat$education)[1]
)

pred_dat$prob_smoke <- predict(
  smoke_model,
  newdata = pred_dat,
  type = "response"
)

ggplot(pred_dat, aes(x = ego_size, y = prob_smoke)) +
  geom_line(linewidth = 1, color = "steelblue") +
  labs(
    x = "Ego Network Size",
    y = "Predicted Probability of Smoking"
  ) +
  theme_minimal()
```

## Predicting drinking frequency - Poisson Regression

> Are respondents with larger, more diverse, or more kin-centered personal networks more likely to drink alcohol more frequently?

### Outcome variable

`drinking_days` coded as

- 0 = less than once a week
- 8 = less than one drink per week
- 9 = does not drink at all
- 1-7 = days per week

8 and 9 will be recoded as 0. Because they are counts, a Poisson regression model is used.

```{r}
analysis_dat <-  analysis_dat %>% 
  mutate(
    drinking_days_rec = if_else(drinking_days %in% c(8,9), 0, drinking_days))

table(analysis_dat$drinking_days_rec, useNA = "ifany")
```

### Predictors

As before

### Estimating the Model

```{r}
drink_model <- glm(
  drinking_days_rec ~
    ego_size + prop_family + gender_entropy + age + gender + education,
  data = analysis_dat,
  family = poisson,
  na.action = na.omit
)

summary(drink_model)
```

### Interpretation of Model Output

`ego_size` shows a positive, statistically significant association with drinking frequency.

```{r}
exp(0.0135)
```

Each additional alter is associated with a roughly 1.4% increase in number of drinking days per week.

`prop_family` shows negative, statistically significant association with drinking frequency

```{r}
exp(-0.282)
```

Each additional family member is associated with a nearly 25% decrease in number of days.

Gender and age are also statistically significant.

```{r}
exp(coef(drink_model))
```

### Predicted Counts

To visualize the effects of the substantively important predictors, we generate predicted counts while varying one predictor at a time and holding the remaining variables constant at typical values. Here, we focus on the three predictors that showed the clearest associations with drinking frequency in the model: `ego_size`, `prop_family`, and `gender`.

```{r}
size_seq <- seq(
  min(analysis_dat$ego_size, na.rm = T),
  max(analysis_dat$ego_size, na.rm = T),
  length.out = 50
)

family_seq <- seq(0, 1, length.out = 50)

pred_size <- tibble(
  ego_size = size_seq,
  prop_family = mean(analysis_dat$prop_family, na.rm = T),
  gender_entropy = mean(analysis_dat$gender_entropy, na.rm = T),
  age = mean(analysis_dat$age, na.rm = T),
  gender = 0,
  education = levels(analysis_dat$education)[1]
)

pred_size <- pred_size %>% 
  mutate(
    pred = predict(drink_model, newdata = pred_size, type = "response"),
    variable = "ego_size",
    x = ego_size
  )

pred_family <- tibble(
  ego_size = mean(analysis_dat$ego_size, na.rm = TRUE),
  prop_family = family_seq,
  gender_entropy = mean(analysis_dat$gender_entropy, na.rm = TRUE),
  age = mean(analysis_dat$age, na.rm = TRUE),
  gender = 0,
  education = levels(analysis_dat$education)[1]
)

pred_family <- pred_family %>%
  mutate(
    pred = predict(drink_model, newdata = pred_family, type = "response"),
    variable = "prop_family",
    x = prop_family
  )

pred_gender <- tibble(
  ego_size = mean(analysis_dat$ego_size, na.rm = TRUE),
  prop_family = mean(analysis_dat$prop_family, na.rm = TRUE),
  gender_entropy = mean(analysis_dat$gender_entropy, na.rm = TRUE),
  age = mean(analysis_dat$age, na.rm = TRUE),
  gender = c(0, 1),
  education = levels(analysis_dat$education)[1]
)

pred_gender <- pred_gender %>%
  mutate(
    pred = predict(drink_model,
      newdata = pred_gender,
      type = "response"
    ),
    variable = "gender",
    x = gender
  )

pred_all <- bind_rows(pred_size, pred_family, pred_gender)
```


```{r r4s-ego-5}
ggplot(pred_all, aes(x = x, y = pred)) +
  geom_line(linewidth = 1, color = "steelblue") +
  facet_wrap(~variable, scales = "free_x") +
  labs(
    x = "Predictor value",
    y = "Expected drinking days per week"
  ) +
  theme_minimal()
```

Taken together, the predicted counts reinforce the main results of the Poisson model. Larger ego networks are associated with somewhat more frequent drinking, whereas networks with a higher proportion of family members are associated with less frequent drinking. At the same time, women are predicted to drink less frequently than men. As in the previous examples, these predictions are model-based associations rather than causal effects. In particular, whether alters themselves drink alcohol would likely matter for understanding social influence more directly, but the present dataset does not include alter-level information on drinking behavior.













