---
title: "Entropy Analysis"
format: gfm
---

[Source](https://schochastics.github.io/R4SNA/descriptive/netropy.html)

```{r}
options(paged.print = FALSE)
```

```{r}
#| message: false

libraries <- list(
  "igraph", "netropy", "tidyverse", "patchwork", 
  "zeallot", "ggraph"
)
invisible(lapply(libraries, library, character.only = TRUE))
```

Multivariate entropy analysis provides a flexible, model-free framework for studying the dependence structure among multiple variables describing entities and their interactions. In network settings, this requires pre-processing steps.

# Network Data Representation

To apply these methods, variables must be represented on finite discrete range spaces and aligned to a common domain of observation. In network settings, this typically involves preprocessing steps such as discretization and the careful definition of observational units (e.g., nodes, dyads, or higher-order structures). 

Since entropy analysis requires all variables to be defined on a common domain, these differences must be resolved through appropriate transformations. Vertex variables can be transformed to the edge domain, or vice versa. Edge attributes can be constructed directly from vertex attributes.

These transformations involve trade-offs. Extending vertex variables to the edge domain increases the number of observations but also enlarges the joint outcome space. Aggregating edge variables to the vertex domain simplifies the data structure but may reduce variability and finer relational detail. Nevertheless, these transformations are essential because they make it possible to bring variables from different domains into a common analytical framework. The same principles extend naturally to higher-order structures such as triads.

Once variables have been aligned to a common domain and appropriately discretized, the data are ready for multivariate entropy analysis. Entropy can be used for dimensionality reduction by identifying redundant variables and selecting informative subsets. Mutual information, derived from joint entropy, quantifies the strength and form of dependence between variables, including higher-order interactions. Expected conditional entropies can be used to assess predictive power by measuring how much uncertainty about one variable is reduced given knowledge of others. Together, these measures provide a comprehensive approach to understanding association, dependence, and information flow in network data. In the following sections, we illustrate this framework through an application.

# Example - Corporate Law Firm

The network consists of 71 lawyers connected through three types of informal interactions: co-working (undirected), advice-seeking (directed), and friendship (directed). In addition, each lawyer is described by several attributes, including status (associate or partner), gender, office location (Boston, Hartford, or Providence), years at the firm, age, and area of practice (corporate or litigation).

Extract the adjacency matrices and the node attributes.

```{r}
adj_advice <- lawdata[[1]]
adj_friend <- lawdata[[2]]
adj_cowork <- lawdata[[3]]
df_att <- lawdata[[4]]
```

Create new dyad variables for mutual friendship and mutual advice.

```{r}
adj_mutadv <- adj_advice * t(adj_advice)
adj_mutfrn <- adj_friend * t(adj_friend)
```

Create likely friendships, which is an undirected network and does not require reciprocal reporting of friendships,by creating the directed graph object and then extracting the adjacency matrix from the graph assuming it is undirected instead.

```{r}
g_frn <- graph_from_adjacency_matrix(adj_friend, mode = "directed")
g_likfrn <- as_undirected(g_frn)
adj_likfrn <- as_adjacency_matrix(g_likfrn, sparse = FALSE)
```

## Data Editing

### Vertex Variables

```{r}
head(df_att, 5)
```

All variables are categorical except `years` and `age`, which must be transformed to ordinal variables. The empirical cumulative distribution functions can be used to identify cutpoints that produce approximately equally sized groups.

We begin by constructing frequency tables for `years` and `age`, and then compute the corresponding relative and cumulative relative frequencies.

```{r}
tab_years <- table(df_att$years)
frq_years <- tibble(
  years = as.numeric(names(tab_years)),
  freq = as.integer(tab_years)
) %>%
  mutate(
    rel.freq = freq / sum(freq),
    cum.rel.freq = cumsum(rel.freq)
  )

tab_age <- table(df_att$age)
frq_age <- tibble(
  age = as.numeric(names(tab_age)),
  freq = as.integer(tab_age)
) %>%
  mutate(
    rel.freq = freq / sum(freq),
    cum.rel.freq = cumsum(rel.freq)
  )


```

```{r}
#| label: fig-cdfs
#| fig-height: 5
#| fig-width: 10
#| fig-cap: "Empirical cumulative distribution functions (CDFs) for `years at firm` and `age`. The dashed horizontal lines indicate the target quantiles (approximately one-third and two-thirds), while the dotted vertical lines show the selected cutpoints used to define three approximately equally sized categories for each variable."

pyears <- ggplot(frq_years, aes(x = years, y = cum.rel.freq)) +
  geom_step(linewidth = 1) +
  geom_point() +
  geom_hline(yintercept = c(1 / 3, 2 / 3), linetype = "dashed") +
  geom_vline(xintercept = c(3, 13), linetype = "dotted") +
  labs(
    title = "CDF of Years at Firm",
    x = "Years at firm",
    y = "Cumulative relative frequency"
  ) +
  theme_minimal()

page <- ggplot(frq_age, aes(x = age, y = cum.rel.freq)) +
  geom_step(linewidth = 1) +
  geom_point() +
  geom_hline(yintercept = c(1 / 3, 2 / 3), linetype = "dashed") +
  geom_vline(xintercept = c(35, 45), linetype = "dotted") +
  labs(
    title = "CDF of Age",
    x = "Age",
    y = "Cumulative relative frequency"
  ) +
  theme_minimal()

pyears + page
```

We use these plots to choose cutpoints that divide the data into three approximately equally sized groups. For years, the selected cutpoints are 3 and 13, while for age, the selected cutpoints are 35 and 45.

```{r}
att_var <- tibble(
  status = df_att$status,
  gender = df_att$gender,
  office = df_att$office - 1,
  years = if_else(df_att$years <= 3, 0, if_else(df_att$years <= 13, 1, 2)),
  age = if_else(df_att$age <= 35, 0, if_else(df_att$age <= 45, 1, 2)),
  practice = df_att$practice,
  lawschool = df_att$lawschool - 1
)
summary(att_var)
```

All variables except years and age are already categorical with finite range spaces and are therefore retained in their original form, apart from simple recoding to start at where needed. The variables years and age are recoded into three ordinal categories.

Finally, we examine the resulting categorized distributions for years and age to confirm that the selected cutpoints produce reasonably balanced groups. 

```{r}
#| label: fig-cats
#| fig-height: 5
#| fig-width: 10
#| fig-cap: "Frequency distributions of the discretized variables `age` and `years`, showing approximately balanced categories after CDF-based grouping."

years_cat <- as.data.frame(table(att_var$years))
names(years_cat) <- c("category", "freq")
years_cat$variable <- "years"

age_cat <- as.data.frame(table(att_var$age))
names(age_cat) <- c("category", "freq")
age_cat$variable <- "age"

att_long <- rbind(years_cat, age_cat)

ggplot(att_long, aes(x = factor(category), y = freq)) +
  geom_col(fill = "lightgrey", color = "darkgrey") +
  facet_wrap(~variable, scales = "free_y") +
  labs(
    x = "Category",
    y = "Frequency"
  ) +
  theme_minimal()
```

Next, we transform the observed dyad variables into node-level variables. This is achieved by aggregating edge information into node-based measures. In particular, we use degree-based summaries: in-degree and out-degree for the directed advice and friendship networks, and degree (number of ties) for the undirected co-work network, as well as for the mutual advice, mutual friendship, and likely friendship networks.

```{r}
g_cwk <- graph_from_adjacency_matrix(
  adj_cowork, mode = "undirected"
)
g_adv <- graph_from_adjacency_matrix(
  adj_advice, mode = "directed"
)
g_mutadv <- graph_from_adjacency_matrix(
  adj_mutadv, mode = "undirected"
)
g_mutfrn <- graph_from_adjacency_matrix(
  adj_mutfrn, mode = "undirected"
)
```

```{r}
deg_cwk <- degree(g_cwk)
indeg_adv <- degree(g_adv, mode = "in")
outdeg_adv <- degree(g_adv, mode = "out")
deg_mutadv <- degree(g_mutadv)
indeg_frn <- degree(g_frn, mode = "in")
outdeg_frn <- degree(g_frn, mode = "out")
deg_mutfrn <- degree(g_mutfrn)
deg_likfrn <- degree(g_likfrn)
```

```{r}
deg_var <- data.frame(
  deg_cwk = ifelse(deg_cwk <= 9, 0, 1),
  indeg_adv = ifelse(indeg_adv <= 10, 0, 1),
  outdeg_adv = ifelse(outdeg_adv <= 11, 0, 1),
  deg_mutadv = ifelse(deg_mutadv <= 3, 0, 1),
  indeg_frn = ifelse(indeg_frn <= 7, 0, 1),
  outdeg_frn = ifelse(outdeg_frn <= 7, 0, 1),
  deg_mutfrn = ifelse(deg_mutfrn <= 4, 0, 1),
  deg_likfrn = ifelse(deg_likfrn <= 11, 0, 1)
)
```

Merging deg_var and att_var yields the final data frame vertex_var, which contains all 15 observed and derived vertex variables used in the subsequent entropy analysis.

```{r}
vertex_var <- cbind(att_var, deg_var)
```

```{r}
head(vertex_var)
```

### Dyad variables

Constructed from pairs of vertex variables, resulting in $\binom{71}{2} = 2485$ observations. `status` yields four possible outcomes: $(0,1),(0,1),(1,0),(1,1)$ and `office` gives nine: $(0,1),(0,1),(0,2),(1,0),(1,1),(1,2),(2,0),(2,1),(2,2)$.

```{r}
c(
  dyad_status, dyad_gender, dyad_office, dyad_years,
  dyad_age, dyad_practice, dyad_lawschool
) %<-%
  map(
    list(
      att_var$status, att_var$gender, att_var$office,
      att_var$years, att_var$age, att_var$practice, att_var$lawschool
    ),
    \(x) get_dyad_var(x, type = "att")
  )
```

Dyadic outcomes are recoded to numeric values, eg. `status` takes values 0-3 and `office` takes values 0-8.

To construct variables based on ties, provide the adjacency matrix.

```{r}
c(dyad_cwk, dyad_adv, dyad_mutadv, 
  dyad_frn, dyad_mutfrn, dyad_likfrn) %<-%
  map(list(adj_cowork, adj_advice, adj_mutadv,
           adj_friend, adj_mutfrn, adj_likfrn),
      \(x) get_dyad_var(x, type = "tie"))
```
Combine all variables.

```{r}
dyad_var <- data.frame(
  status = dyad_status$var,
  gender = dyad_gender$var,
  office = dyad_office$var,
  years = dyad_years$var,
  age = dyad_age$var,
  practice = dyad_practice$var,
  lawschool = dyad_lawschool$var,
  cowork = dyad_cwk$var,
  advice = dyad_adv$var,
  mutadv = dyad_mutadv$var,
  friend = dyad_frn$var,
  mutfrn = dyad_mutfrn$var,
  likfrn = dyad_likfrn$var
)
head(dyad_var, 10)
```

### Triad variables

Repead the process. The combinations of three nodes gives $\binom{71}{3} = 57155$ observations.

```{r}
c(triad_status, triad_gender, triad_office, triad_years,
  triad_age, triad_practice, triad_lawschool) %<-%
  map(list(att_var$status, att_var$gender, att_var$office, att_var$years,
           att_var$age, att_var$practice, att_var$lawschool),
      \(x) get_triad_var(x, type = "att"))
```

```{r}
c(triad_cwk, triad_adv, triad_mutadv, 
  triad_frn, triad_mutfrn, triad_likfrn) %<-%
  map(list(adj_cowork, adj_advice, adj_mutadv,
           adj_friend, adj_mutfrn, adj_likfrn),
      \(x) get_triad_var(x, type = "tie"))
```

```{r}
triad_var <- data.frame(
  status = triad_status$var,
  gender = triad_gender$var,
  office = triad_office$var,
  years = triad_years$var,
  age = triad_age$var,
  practice = triad_practice$var,
  lawschool = triad_lawschool$var,
  cowork = triad_cwk$var,
  advice = triad_adv$var,
  mutadv = triad_mutadv$var,
  friend = triad_frn$var,
  mutfrn = triad_mutfrn$var,
  likfrn = triad_likfrn$var
)
head(triad_var, 10)
```

## Univariate and Bivariate Entropies

Univariate entropy measures the uncertainty of a discrete variable $X$ with $r$ outcomes. Entropy of 0 means no uncertainty, ie. it is constant. The maximum is $\log_2 r$ corresponding to a uniform distribution over all outcomes. It is defined:

$$
H(X) = \sum_x p(x) \log_2 \frac{1}{p(x)}.
$$

Bivariate entropy assesses redundancy, functional relationships and stochastic independence. It is given

$$
H(X,Y) = \sum_x \sum_y p(x,y) \log_2 \frac{1}{p(x,y)}.
$$

The bounds are

$$
H(X) \leq H(X,Y) \leq H(X) + H(Y).
$$

- The lower bound is attained if and only if there is a functional relationship $Y = f(X)$.
- The upper bound is attained if and only if $X$ and $Y$ are stochastically independent, $X \perp Y$.
- If $H(X,Y) = H(X)$ or $H(X,Y) = H(Y)$, one of the variables is redundant and does not provide additional information.

The function entropy_bivar() computes the bivariate entropies for all pairs of variables in a data frame. The results are returned as an upper triangular matrix, where each cell contains the joint entropy of the corresponding row and column variables. The diagonal entries represent the univariate entropies of each variable.

### Vertex variables

```{r}
h2_vertex <- entropy_bivar(vertex_var)
h2_vertex[1:8,1:8]
```

Redundancy can be identified by comparing the diagonal, univariate entropies with the other values in the same row. If they are equal, it indicates one variable is redundant. Here none are. Can check `redundancy`. `dec` controls rounding of bivariate entropies.

```{r}
(red_vertex <- redundancy(vertex_var, dec = 3))
```

```{r}
(red_vertex2 <- redundancy(df_att, dec = 2))
```

Senior consisted entirely of unique values and is flagged as redundant. To identify redundant variables:

```{r}
colnames(red_vertex2)[rowSums(red_vertex2) >= 1]
```

### Dyad variables

```{r}
(h2_dyad <- entropy_bivar(dyad_var))
```

NB. this does not work on tibbles.

Redundancies can be seen for `friend` and `advice` The variables `mutadv`, `mutfrn` and `likfrn` should be dropped.

```{r}
dyad_var <- dyad_var[-c(10, 12:13)]
```

### Triad variables

```{r}
(h2_triad <- entropy_bivar(triad_var))
```

The same redundancy are evident, and are removed.

```{r}
triad_var <- triad_var[-c(10, 12:13)]
```

## Joint Entropies and Association Graphs

Measures the total uncertainty associated with a pair of variables. Provides a basis for assessing their dependence.

$$
J(X,Y) = H(X) + H(Y) - H(X,Y)
$$

It equals 0 if and only if $X$ and $Y$ are stochastically independent, $X \perp Y$.

For three variables, trivariate entropy is bound by

$$
H(X,Y) \leq H(X,Y,Z) \leq H(X,Z) + H(Y,Z) - H(Z)
$$

and deviation from the upper bound

$$
EJ(X,Y \mid Z) = H(X,Z) + H(Y,Z) - H(Z) - H(X,Y,Z)
$$

which is 0 if and only if $X \perp Y \mid Z$.

Joint entropy values can be used to explore dependence structure through association graphs, with variables as nodes and edges connecting pairs whose entropy exceeds a chosen threshold. Lowering the threshold gives more edges.

Cliques indicate groups of mutually dependent variables. Conditional independence can by investigated by examining how the graph structure changes when conditioning variables are removed. 

The function `joint_entropy()` computes joint entropy values for all pairs of variables in a data frame. The output is returned as a list containing: (i) an upper triangular matrix of joint entropy values, with univariate entropies on the diagonal, and (ii) a data frame summarizing the frequency distribution of unique joint entropy values. An optional argument controls the precision (number of decimal places) used in constructing this frequency distribution; the default is 3.

### Vertex Variables

```{r}
j_vertex <- joint_entropy(vertex_var, 3)
str(j_vertex)
```

```{r}
j_vertex$matrix
```

The strongest associations can be identified by combining the matrix with the frequency table `j_vertex$freq`, which summarizes the distribution of joint entropy values. 

```{r}
j_vertex$freq
```

```{r r4sn-ent-1}
assoc_graph(vertex_var, 0.25)
```

### Dyad variables

Same.

```{r r4sn-ent-2}
assoc_graph(dyad_var, 0.10)
```

There is a strong dependence among the variables (status, years, age), as well as notable associations among (lawschool, years, age) and (status, years, gender).

In addition, the graph suggests that the relations cowork and friend are conditionally independent given advice. That is, any observed dependence between cowork and friend can be explained through their association with advice.

### Triad Variables

Same

```{r r4sn-ent-3}
assoc_graph(triad_var, 0.25)
```

The triad variables exhibit similar dependence patterns as the dyad variables, but with generally stronger associations. In particular, joint entropy values are higher for all pairs of cowork, friend, and advice, indicating stronger interactions among these variables. This further supports the conditional independence observed earlier, suggesting that the relationship between cowork and friend is increasingly explained by advice at the triadic level.

## Trivariate Entropies and Prediction Power

Similar to bivariate entropy.

$$
H(X,Y,Z) = \sum_x \sum_y \sum_z p(x,y,z) \log_2 \frac{1}{p(x,y,z)},
$$
with bounds

$$
H(X,Y) \leq H(X,Y,Z) \leq H(X,Z) + H(Y,Z) - H(Z).
$$

The deviation from the lower bound is given by the expected conditional entropy

$$
EH(Z \mid X,Y) = H(X,Y,Z) - H(X,Y),
$$
which measures the remaining uncertainty in $Z$ given $(X,Y)$. This quantity is non-negative and equals 0 if and only if $(X,Y)$ uniquely determine $Z$, indicating a functional relationship.

More generally, $EH(Z \mid X,Y)$ can be interpreted as a measure of prediction uncertainty when $(X,Y)$ are used to predict $Z$. It reflects, on a logarithmic scale, the average number of possible outcomes of $Z$ given $(X,Y)$. Smaller values indicate stronger predictive power: if $EH < 0.5$, $Z$ can be predicted unambiguously; if $0.5 \leq EH < 1.5$, there are approximately two possible outcomes on average; and so on. In general, prediction power decreases as $EH$ increases.

The function `prediction_power()` computes this quantity for all pairs of variables in a given data frame used to predict a specified target variable $Z$. The function takes as input the data frame and the variable to be predicted, and returns an upper triangular matrix of expected conditional entropies $EH(Z \mid X,Y)$ for all pairs $(X,Y)$. The diagonal entries correspond to $EH(Z \mid X)$, that is, prediction based on a single variable. Entries corresponding to the target variable itself are set to `NA`.

For easier interpretation, prediction results can be visualized using prediction plots, which display a color-coded matrix with rows representing $X$ and columns representing $Y$. Darker colors indicate lower prediction uncertainty and thus higher predictive power for $Z$. In the following, we compute and visualize prediction power for vertex, dyad, and triad variables.

### Vertex Variables

```{r}
h3_vertex <- entropy_trivar(vertex_var)
```

```{r}
c(
  pred_vertex_cwk, pred_vertex_advin, pred_vertex_advout,
  pred_vertex_advmut, pred_vertex_frnin, pred_vertex_frnout,
  pred_vertex_frnmut, pred_vertex_frnlik
) %<-%
  map(
    list(
      "deg_cwk", "indeg_adv", "outdeg_adv",
      "deg_mutadv", "indeg_frn", "outdeg_frn",
      "deg_mutfrn", "deg_likfrn"
    ),
    \(x) prediction_power(x, vertex_var)
  )
```

As an illustration, consider predicting the variable `indeg_adv`, which represents the in-degree in the advice network (i.e., how often a lawyer is asked for advice). The prediction power was computed and assigned to `pred_vertex_advin`.

```{r}
pred_vertex_advin
```

```{r}
#| label: fig-expred
#| fig-height: 7
#| fig-width: 8
#| fig-cap: "Prediction power heatmap for $Z$=`indeg_adv`. Each cell shows the expected conditional entropy $EH(Z \\mid X,Y)$, where darker colors indicate stronger predictive power. The strongest predictors involve the variable `years`, particularly in combination with network-based variables such as `deg_mutadv` and `deg_cwk`. While `status` also contributes to prediction, it is not part of the strongest predictor pairs."

make_pred_plot(pred_vertex_advin, "indeg_adv")
```

`deg_likfrn` is the strongest overall predictor of `indeg_adv`. Pairwise, `years` with the network variables `deg_mutadv` and `deg_cwk` are strong.
These results suggest that both network position, particularly likely friendship ties, and experience (years) play a central role in explaining variation in advice in-degree.

```{r}
#| label: fig-predplot-vertex
#| fig-height: 18
#| fig-width: 10
#| fig-cap: "Prediction power heatmaps for vertex variables; darker cells indicate stronger predictive power (lower $EH(Z \\mid X,Y)$)."

c(p1, p2, p3, p4, p5, p6, p7, p8) %<-%
  map2(
    list(
      pred_vertex_cwk, pred_vertex_advin, pred_vertex_advout,
      pred_vertex_advmut, pred_vertex_frnin, pred_vertex_frnout,
      pred_vertex_frnmut, pred_vertex_frnlik
    ),
    list(
      "deg_cwk", "indeg_adv", "outdeg_adv",
      "deg_mutadv", "indeg_frn", "outdeg_frn",
      "deg_mutfrn", "deg_likfrn"
    ),
    \(x, y) make_pred_plot(x, y)
  )

(p1 | p2) /
  (p3 | p4) /
  (p5 | p6) /
  (p7 | p8)
```

The vertex variables, network-based measures such as deg_likfrn, deg_mutadv, and other degree-related variables are consistently strong predictors, both individually and in combination. In particular, deg_likfrn often emerges as the best single predictor, indicating that embeddedness in likely friendship ties plays a central role in explaining variation in network position.

Attribute-based variables such as years and age also contribute to prediction, especially when combined with network measures, while variables such as office and practice tend to have weaker predictive power. Overall, the results highlight that structural position in the network is more informative than individual attributes for predicting vertex-level outcomes.

### Dyad variables

Assess the prediction power for all dyad variables. For each dyad variable, compute the expected conditional entropies $EH(Z \mid X,Y)$ using all other dyad variables as predictors.

```{r}
var_list <- list(
  "status", "gender", "office", "years", "age",
  "practice", "lawschool", "cowork", "advice", "friend"
)
c(
  pred_dyad_status, pred_dyad_gender, pred_dyad_office,
  pred_dyad_years, pred_dyad_age, pred_dyad_practice,
  pred_dyad_lawschool, pred_dyad_cowork, pred_dyad_advice,
  pred_dyad_friend
) %<-%
  map(
    var_list,
    \(x) prediction_power(x, dyad_var)
  )
```

```{r}
#| label: fig-predplot-dyad
#| fig-height: 18
#| fig-width: 10
#| fig-cap: "Prediction power heatmaps for dyad variables; darker cells indicate stronger predictive power (lower $EH(Z \\mid X,Y)$)."

c(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10) %<-%
  map2(
    list(
      pred_dyad_status, pred_dyad_gender, pred_dyad_office,
      pred_dyad_years, pred_dyad_age, pred_dyad_practice,
      pred_dyad_lawschool, pred_dyad_cowork, pred_dyad_advice,
      pred_dyad_friend
    ),
    var_list,
    \(x, y) make_pred_plot(x, y)
  )

(p1 | p2) /
  (p3 | p4) /
  (p5 | p6) /
  (p7 | p8) /
  (p9 | p10)
```

Overall, the plots show that relational variables such as advice, cowork, and friend are strong predictors of each other, particularly when combined. Attribute-based variables such as years and age contribute to prediction, but are generally less informative on their own, highlighting the importance of network structure in explaining dyadic interactions.

### Triad variables

```{r}
var_list[1:7]
```


```{r}
#| label: fig-predplot-triad
#| fig-height: 18
#| fig-width: 10
#| fig-cap: "Prediction power heatmaps for triad variables; darker cells indicate stronger predictive power (lower $EH(Z \\mid X,Y)$)."

var_list <- var_list[1:7]

# compute prediction power matrices
c(pred_triad_status, pred_triad_gender, pred_triad_office,
  pred_triad_years, pred_triad_age, pred_triad_practice,
  pred_triad_lawschool) %<-% 
  map(var_list,
      \(x) prediction_power(x, triad_var))

c(p1, p2, p3, p4, p5, p6, p7) %<-%
  map2(
    list(
      pred_triad_status, pred_triad_gender, pred_triad_office,
      pred_triad_years, pred_triad_age, pred_triad_practice,
      pred_triad_lawschool
    ),
    var_list,
    \(x, y) make_pred_plot(x, y)
  )
(p1 | p2) /
  (p3 | p4) /
  (p5 | p6) /
  (p7 | patchwork::plot_spacer())
```

For the triadic attribute variables, status, years, and gender are predicted relatively well, with especially low expected conditional entropy for status and years. In particular, status is most informative for predicting both years and gender, while combinations involving office and years are especially useful for predicting status. By contrast, office, practice, and lawschool are much harder to predict, with generally higher expected conditional entropy values across the corresponding heatmaps.

Some of the clearest gains from using pairs of predictors appear for office, where combinations involving structural variables markedly reduce prediction uncertainty. In particular, the pair (cowork, advice) provides a very strong prediction of triadic office, much stronger than any single predictor alone. For gender, prediction is moderate overall, with status serving as the best single predictor.

The triad-level results suggest that attribute-based variables such as status, years, and gender are often the most informative single predictors, while structural variables such as cowork and advice become especially important when used jointly. Thus, the triadic representation captures richer interactions, but the strongest predictive relationships arise from specific combinations of attribute and structural information rather than from structural variables alone.

## Divergence Tests of Goodness of Fit

Compare the _empirical distribution_ $p(x)$ to a _model distribution_ $q(x)$ representing a hypothesized structure.

The discrepancy is measured by _information divergence_

$$
D(p,q) = \sum_x p(x)\log \frac{p(x)}{q(x)}
$$

which is the expected log-likelihood ratio between the models. The test statistic is

$$
2n D(p,q),
$$

which, for large samples, is approximately $\chi^2$-distributed with degrees of freedom determined by the difference in model complexity.

#### Independence testing

##### Pairwise independence

To test independence of $X$ and $Y$, compare the empirical joint distribution $p(x,y)$ and the model $q(x,y) = p(x)p(y)$. The divergence is

$$
D(p,q) = H(X) + H(Y) - H(X,Y) = J(X,Y),
$$

and the test statistic is $2n J(X,Y)$.

Thus, joint entropy directly measures deviation from independence: small values indicate independence, while large values indicate dependence.

##### Conditional independence

Are two variables independent given a third variable, so $X \perp Y \mid Z$. The model distribution is

$$
q(x,y,z) = \frac{p(x,z)p(y,z)}{p(z)},
$$

and the divergence is

$$
D(p,q) = H(X,Z) + H(Y,Z) - H(Z) - H(X,Y,Z) = EJ(X,Y \mid Z),
$$

And the test statistic is $2n\,EJ(X,Y \mid Z)$

##### Nested model specifications

Compare models where one is a constrained version of the other. If the model $q_2$ is nested in $q_1$, ie. it imposes additional assumptions, the difference in divergence can be used to test if the additional constraints are supported by the data.

$$
2n\bigl[D(p,q_2) - D(p,q_1)\bigr]
$$

#### Example - dyad variables

The association graph constructed above suggests candidate dependence structures that can be formally evaluated. In the corporate law firm example, the variables `friend`, `cowork`, and `advice` form a closely connected group, motivating the conditional independence model

```{r}
#| label: fig-divtest
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Association graph for dyad variables (threshold $J > 0.1$). The variables `friend`, `cowork`, and `advice` are highlighted, motivating the conditional independence model $\\text{friend} \\perp \\text{cowork} \\mid \\text{advice}$."

p <- assoc_graph(dyad_var, 0.10)

p +
  geom_node_point(
    data = p$data,
    aes(x = x, y = y, fill = name %in% c("friend", "cowork", "advice")),
    shape = 21,
    color = "black",
    size = 20
  ) +
  scale_fill_manual(
    values = c("FALSE" = "white", "TRUE" = "tan1"),
    guide = "none"
  ) +
  geom_node_text(
    data = p$data,
    aes(x = x, y = y, label = name),
    color = "black"
  )
```

Intuitively, this model says that any association between friendship and co-working may be explained by the advice relation. In other words, once `advice` is taken into account, there should be little additional dependence between `friend` and `cowork`. To test

```{r}
div_gof(
  dat = dyad_var,
  var1 = "friend",
  var2 = "cowork",
  var_cond = "advice"
)
```

To test $\text{gender} \perp \text{practice} \mid \text{status}$

```{r}
div_gof(
  dat = dyad_var,
  var1 = "gender",
  var2 = "practice",
  var_cond = "status"
)
```

To compare, test without conditioning.

```{r}
div_gof(
  dat = dyad_var, var1 = "friend", var2 = "cowork"
)
```

Can also check if uniformly distributed

```{r}
div_gof(dat = dyad_var, var_uniform = "friend")
```

Finally, reduced models can be compared to the saturated empirical model. The saturated model corresponds to the unrestricted empirical distribution and has divergence equal to zero. For example, we can compare the pairwise independence model for `friend` and `cowork` to the saturated model.

```{r}
m_full <- list(D = 0, df = 0)
m_reduced <- div_gof(dat = dyad_var, var1 = "friend", var2 = "cowork")

div_gof(
  dat = dyad_var,
  model_full = m_full,
  model_reduced = list(D = m_reduced$D, df = m_reduced$df)
)
```

The divergence-based test statistic used in `div_gof()` is

$$
\chi^2 = 2nD \cdot \log(2),
$$

which is approximately chi-square distributed with degrees of freedom depending on the specified model.

By default, the function uses a large-sample approximation to the 5% critical value,

$$
\text{cv} \approx df + \sqrt{8\,df},
$$

which provides a fast and convenient rule for hypothesis testing.

Alternatively, exact critical values from the chi-square distribution can be used by setting.

```{r}
#| eval: false
use_approx_cv = FALSE
```

In this case, the significance level can be controlled explicitly via the argument alpha (default is 0.05), allowing for more precise hypothesis testing.

```{r}
#| eval: false
div_gof(..., use_approx_cv = FALSE, alpha = 0.01)
```

Lower values of alpha correspond to stricter tests (harder to reject the model), while higher values make rejection more likely. In particular, smaller alpha reduces the probability of incorrectly rejecting a true null hypothesis.

