---
title: "Enhancing Visualizations"
format: gfm
---

[Source](https://schochastics.github.io/R4SNA/visualization/interactive-viz.html)

```{r}
options(paged.print = FALSE)
```

```{r}
#| message: false

libraries <- list(
  "igraph", "visNetwork", "networkD3",
  "threejs", "networkdata", "g6R", "tidyverse"
)
invisible(lapply(libraries, library, character.only = TRUE))
```

# Data Preparation

We use the Florentine Families marriage network, introduced in earlier chapters. With 16 families and 20 ties it is small enough that every label, hull, and bubble in this chapter renders without clutter, and it comes with a wealth attribute that gives us a meaningful categorical grouping for later plugin demos. The disconnected Pucci family will be removed.

```{r}
flo <- subgraph(
  flo_marriage,
  components(flo_marriage)$membership ==
    which.max(components(flo_marriage)$csize)
)
```

Assign wealth tiers for grouping and degree for node sizes.

```{r}
V(flo)$wealth_tier <- as.character(
  cut(
    V(flo)$wealth,
    breaks = quantile(V(flo)$wealth, probs = c(0, 1/3, 2/3, 1)),
    labels = c("low", "mid", "high"),
    include.lowest = TRUE
  )
)

V(flo)$degree <- degree(flo)


palette3 <- c(low = "#1A5878", mid = "#AD8941", high = "#C44237")
```

# Interactive Network Visualization Tools

Each of these packages has its place. `visNetwork` is mature and full-featured, `networkD3` leverages the power of `D3.js`, and `threejs` offers 3D rendering. 

## visNetwork

Converting an igraph object requires mapping a few vertex attributes to the column names that visNetwork expects.

```{r}
V(flo)$label <- V(flo)$name
V(flo)$group <- V(flo)$wealth_tier
V(flo)$value <- V(flo)$degree

vis_data <- toVisNetworkData(flo)
visNetwork(
  vis_data$nodes, vis_data$edges,
  width = "100%", height = "400px"
) %>% 
  visOptions(highlightNearest = TRUE)

```

A simple, draggable and zoomable tool.

## networkD3

```{r}
d3_data <- igraph_to_networkD3(flo, group = V(flo)$wealth_tier)

forceNetwork(
  Links = d3_data$links, Nodes = d3_data$nodes,
  Source = "source", Target = "target",
  NodeID = "name", Group = "group",
  opacity = 0.9, zoom = TRUE, fontSize = 12
)
```

Hovering highlights connected nodes and fades others.

## threejs

3D rendering

```{r}
graphjs(flo, vertex.size = 0.3 + 0.2 * V(flo)$degree)
```

# g6R Basics

## Building nodes and edges

`g6R` offers a convenience function `g6_igraph` to convert an `igraph` object to a g6 widget, but it is only good for a quick preview. To fully control, must build the graph explicitly.

The g6 node is a list with at least an `id`, and optional `data` and `style` slots. The edge is a list with `source` and `target` fields.

```{r}
mk_node <- function(name, tier, deg) {
  g6_node(
    id = name,
    data = list(wealth_tier = tier, degree = deg),
    style = list(
      labelText = name,
      fill = palette3[[tier]],
      size = 14 + 2 * deg,
      stroke = "#555555",
      lineWidth = 1
    )
  )
}

nodes <- pmap(
  list(V(flo)$name, V(flo)$wealth_tier, V(flo)$degree),
  \(x, y, z) mk_node(x, y, z)
)
```

```{r}
edges_df <- igraph::as_data_frame(flo, what = "edges")

mk_edge <- function(src, tgt) {
  g6_edge(
    source = src,
    target = tgt,
    style = list(stroke = "#cccccc", lineWidth = 1)
  )
}

edges <- map2(
  edges_df$from,
  edges_df$to,
  \(x, y) mk_edge(x, y)
) %>%
  unname() %>%
  sapply(g6_edges)
```

## First interactive network

```{r}
g6(nodes, edges) %>% 
  g6_layout() %>% 
  g6_options(autoFit = "view", autoResize = TRUE)
```

#  Layouts

The default layout in g6R is d3_force_layout(), which simulates physical forces to position nodes: linked nodes attract each other while all nodes repel, producing a layout that tends to place densely connected groups close together. 

Another force-directed option is force_atlas2_layout(), a variant of the ForceAtlas2 algorithm popularized by Gephi. It produces a more compact layout with less overlap, and the preventOverlap option can be set to TRUE to further reduce node collisions at the cost of a longer layout time. `kr` increases the repulsion strength.

```{r}
g6(nodes, edges) %>% 
  g6_layout(force_atlas2_layout(preventOverlap = T, kr = 20)) %>% 
  g6_options(autoFit = "view", autoResize = T)
```

# Styling nodes and edges

Per-element styles, set in the style slot of each node or edge, give every element its own color, size, and label as we did in the data preparation above. Global defaults, set with g6_options(), apply to all elements and can be overridden by per-element styles.

## Per-element styling

```{r}
mk_node_bold <- function(name, tier, deg) {
  g6_node(
    id = name,
    data = list(wealth_tier = tier, degree = deg),
    style = list(
      labelText = name,
      fill = palette3[[tier]],
      size = 14 + 2 * deg,
      stroke = "#222222",
      lineWidth = 2.5
    )
  )
}

nodes_bold <- pmap(
  list(V(flo)$name, V(flo)$wealth_tier, V(flo)$degree),
  \(x, y, z) mk_node_bold(x, y, z)
)

g6(nodes_bold, edges) %>% 
  g6_layout(force_atlas2_layout(preventOverlap = T, kr = 20)) %>% 
  g6_options(autoFit = "view", autoResize = T)  
```

## Global defaults

Shrink the label font, move labels below nodes.

```{r}
g6(nodes, edges) %>%
  g6_layout(force_atlas2_layout(preventOverlap = T, kr = 20)) %>%
  g6_options(
    node = node_options(
      style = node_style_options(
        labelFontSize = 11,
        labelPlacement = "bottom"
      )
    ),
    edge = edge_options(
      style = edge_style_options(
        stroke = "#bbbbbb",
        lineWidth = 0.8
      )
    ),
    autoFit = "view",
    autoResize = TRUE
  )
```

# Behaviors

## Navigation: drag and zoom

```{r}
g6(nodes, edges) |>
  g6_layout(force_atlas2_layout(preventOverlap = TRUE, kr = 20)) |>
  g6_behaviors(drag_canvas(), zoom_canvas())
```

## Element interaction

```{r}
g6(nodes, edges) %>% 
  g6_layout() %>% 
  g6_behaviors(
    drag_element_force(fixed = TRUE),
    hover_activate(degree = 1),
    click_select(multiple = TRUE)
  )
```

## Lasso and brush selection

```{r}
g6(nodes, edges) |>
  g6_layout() |>
  g6_behaviors(
    lasso_select()
  )
```

```{r}
g6(nodes, edges) |>
  g6_layout() |>
  g6_behaviors(
    brush_select()
  )
```

# Plugins

## Visual grouping: hulls

```{r}
tier_members <- split(V(flo)$name, V(flo)$wealth_tier)

g6(nodes, edges) %>% 
  g6_layout() %>% 
  g6_behaviors("drag-element") %>% 
  g6_options(
    animation = FALSE
  ) %>% 
  g6_plugins(
    hull(
      key = "hull-low", members = tier_members$low,
      fill = palette3[["low"]], stroke = palette3[["low"]]
    ),
    hull(
      key = "hull-mid", members = tier_members$mid,
      fill = palette3[["mid"]], stroke = palette3[["mid"]]
    ),
    hull(
      key = "hull-high", members = tier_members$high,
      fill = palette3[["high"]], stroke = palette3[["high"]]
    )
  )

```

## Visual grouping: bubble sets

```{r}
g6(nodes, edges) %>% 
  g6_layout() %>% 
  g6_behaviors("drag-element") %>% 
  g6_options(
    animation = FALSE
  ) %>% 
  g6_plugins(
    bubble_sets(
      key = "hull-low", members = tier_members$low,
      fill = palette3[["low"]], stroke = palette3[["low"]]
    ),
    bubble_sets(
      key = "hull-mid", members = tier_members$mid,
      fill = palette3[["mid"]], stroke = palette3[["mid"]]
    ),
    bubble_sets(
      key = "hull-high", members = tier_members$high,
      fill = palette3[["high"]], stroke = palette3[["high"]]
    )
  )
```

## Tooltips

The callback is a small JavaScript function, wrapped with JS() that receives the hovered items and returns the HTML to render. 

```{r}
tooltip_content <- JS(
  "(event, items) => {
    let result = ``;
    items.forEach((item) => {
      result += `<h4>${item.id}</h4>`;
      result += `<p>wealth tier: ${item.data.wealth_tier}</p>`;
      result += `<p>degree: ${item.data.degree}</p>`;
    });
    return result;
  }"
)

g6(nodes, edges) |>
  g6_layout() |>
  g6_plugins(tooltips(trigger = "click",getContent = tooltip_content))
```

## Legend

```{r}
g6(nodes, edges) |>
  g6_layout() |>
  g6_plugins(
    legend(
      nodeField = "wealth_tier",
      showTitle = TRUE, 
      titleText = "Wealth Tier", 
      position = "top"
    )
  )
```

## Fisheye lens

```{r}
g6(nodes, edges) |>
  g6_layout() |>
  g6_plugins(fish_eye())
```

## Minimap

```{r}
g6(nodes, edges) |>
  g6_layout() |>
  g6_plugins(minimap())
```




















