---
title: "Stochastic Actor Oriented Models (SAOMs)"
format: gfm
---

[Source](https://schochastics.github.io/R4SNA/inferential/saom.html)

```{r}
options(paged.print = FALSE)
```

```{r}
#| message: false

libraries <- list(
  "igraph", "ggraph","graphlayouts",
  "networkdata", "intergraph", "RSiena",
  "tidyverse", "zeallot"
)
invisible(lapply(libraries, library, character.only = TRUE))
```

Stochastic Actor-Oriented Models (SAOMs) are explicitly designed for longitudinal network data. This allows researchers to address questions such as:

- How often do actors initiate or terminate ties?
- Do actors prefer to form ties based on similarity (selection) or become more similar after forming ties (influence)?
- How do structural tendencies like reciprocity or transitivity manifest through sequential decision-making?
- Can we simulate or predict how the network may evolve in future time points?

**Temporal Exponential Random Graph Models (TERGM)** can also be used, but SOAM's focus on process, how ties change due to actor decisions.

# Running Example: Teenage Friends and Lifestyle Study

We use a running example with data from the "Teenage Friends and Lifestyle Study" [@west1996s50], which was also used in the previous chapter on ERGMs. In contrast to the earlier example that focused on a single time point and only the subset of 50 pupils, we here model the dynamics of friendship ties across all three observed waves and with the 129 pupils over all three time points, and including relevant actor-level covariates such as gender and smoking behavior. The dataset is called `glasgow129` in the `networkdata` package, which includes the following:

- Networks: Binary, directed friendship ties measured at three time points.
- Actors: 129 pupils measured at three time points.
- Many covariates, but we focus on
  - Sex: 1 = Male, 2 = Female
  - Smoking: 1 = No, 2 = Occasional, 3 = Regular

```{r}
c(glasgow_g1, glasgow_g2, glasgow_g3) %<-% 
  map(1:3, \(i) glasgow129[[i]])
c(net1, net2, net3) %<-%
  map(list(glasgow_g1, glasgow_g2, glasgow_g3),
      \(x) as_adjacency_matrix(x, sparse = F))
net_array <- array(c(net1, net2, net3), dim = c(129, 129, 3))
```

# SAOM Framework

## Modeling Network Evolution

Longitudinal network data consist of a set of actors and a series of observed adjacency matrices, each capturing ties at a particular point in time.

These repeated measures make it possible to ask and answer key questions:

- How frequently do actors change their ties?
- What drives the formation, maintenance, or dissolution of ties?
- How do individual attributes (e.g., gender, age, group membership) shape network dynamics?
- Can we predict how the network will evolve in the future?
- How do both endogenous (network-based) and exogenous (attribute-based) factors jointly shape the network?

Stochastic Actor-Oriented Models (SAOMs) offer a solution by treating network change as a **continuous-time, actor-driven process**. In this framework:

- The network evolves through a sequence of micro-steps where individual actors have opportunities to change their outgoing ties.
- Each actor evaluates the current network and makes decisions based on preferences (e.g., for reciprocation, closure, or similarity).
- The model simulates the timing and direction of these changes between observation moments (waves).

# Estimating SAOMs in RSiena

Can model network evolution and co-evolution of networks and behavior. If only a network object is included, it captures how actors form, maintain, or dissolve ties of time. Individual attributes are treated as exogenous and fixed over time. It can explain **social selection**. If a behavioral variable is included, it can also model change of behavior over time, explaining **social influence**.

Below, we outline the typical steps involved in estimating SAOMs using RSiena:

1.  **Preparing the data**\
    Import and format the network and covariate data into the appropriate structure. This includes creating adjacency matrices for networks observed over time and vectors or matrices for actor-level covariates. These are then converted into RSiena objects such as `sienaDependent`, `coCovar`, or `varCovar`.

2.  **Specifying the model**\
    Define the model by selecting the effects to include. Use `getEffects()` to retrieve the default effects object and `includeEffects()` to add endogenous and covariate-related effects.

3.  **Estimating the model**\
    Set up the estimation procedure using `sienaAlgorithmCreate()` and estimate the model with `siena07()`. Evaluate convergence and diagnostic statistics to ensure reliable results.

4.  **Interpreting the results**\
    Analyze the estimated parameters to understand the underlying social processes. This involves examining the direction, magnitude, and statistical significance of effects to interpret mechanisms of network change or co-evolution.


## Model 1: Structural Network Effects

The model includes outdegree and reciprocity effects, capturing a baseline propensity for ties and mutual ties. These are endogenous network dynamics.

Convert 3D network array to a `sienaDependent` object.

```{r}
siena_net <- sienaDependent(net_array)
```

Wrap that into a complete `RSiena` object.

```{r}
#| message: false
data_mod1 <- sienaDataCreate(siena_net)
```

List all possible effects available for the data structure.

```{r}
#| message: false
(my_effects <- getEffects(data_mod1))
```

Define the estimation algorithm. The `projname` argument specifies a name for the project; RSiena will use this name to organize temporary output files and logs related to the estimation.

```{r}
#| message: false
#| output: false
my_algorithm <- sienaAlgorithmCreate(projname = "saom_1", seed = 1108)
```

This algorithm object is later passed into the siena07() function to control how the estimation is performed. You can customize various settings like number of iterations, convergence thresholds, or use defaults (as we do here) for a basic estimation.

The siena07() function. This function runs the RSiena estimation procedure. It takes the following inputs:

- my_algorithm: the algorithm settings we defined earlier.
- data = data_mod1: the network data in RSiena format.
- effects = my_effects: the specified model effects (in this case, outdegree and reciprocity).
- batch = TRUE: suppresses user prompts during the estimation, making it suitable for scripted or automated runs.

```{r}
#| message: false
#| output: false
#| cache: true

saom_1 <- siena07(
  my_algorithm,
  data = data_mod1,
  effects = my_effects,
  batch = TRUE, 
  useCluster = T, nbrNodes = 6
)
```

```{r}
saom_1
```


```{r}
significance <- function(mod) {
  data.frame(
    effect = mod$effects$effectName,
    significance = abs(mod$theta / mod$se),
    gt.20 = ifelse(abs(mod$theta / mod$se) > 2, "Yes", "No")
  )
}
significance((saom_1))
```


### Convergence assessment

Convergence is assessed using two main criteria

1. Overall maximum convergence ratio. Values below 0.25 are considered acceptable, below 0.15 very good
2. Individual convergence t-ratios. Below 0.20 is generally exceptable, below 0.10 are ideal

In this case, the diagnostics confirm that the estimation process stabilized.

### Rate parameters

- Rate 1: On average, each actor had 8.6 opportunities to change ties between waves 1 and 2
- Rate 2: On average, each actor had 7.2 opportunities to change ties between waves 2 and 3

### Network dynamics

- Outdegree density - strong negative effect - actors are reluctant to form many ties and prefer sparser networks. Forming new ties carries a cost 
- Reiprocity - strong positive effect - actors prefer to reciprocate existing ties

The objective function for actor $i$ can be written:

$$
f_i(\theta, Y^{(a)}) =
\theta_{\text{out}} \sum_{j} y^{(a)}_{ij}
+
\theta_{\text{rec}} \sum_{j} y^{(a)}_{ij} \, y^{(a)}_{ji}
$$

To understand this, consider two scenarios:

1. **Adding a reciprocated tie**\
  Suppose $y_{ji}(t) = 1$ (the other actor has already sent a tie), and actor $i$ considers forming $y_{ij} = 1$. The change in utility is -2.415 + 2.711 = 0.296. This yields a **positive utility change**, making the move likely.

2. **Adding a non-reciprocated tie**\
  If $y_{ji}(t) = 0$, the tie is not reciprocated. The utility change is -2.411. This is **negative**, making the change unlikely.

## Model 2: Structural and covariate effects

Incorporating covariates 

into the objective function requires distinguishing between different types of effects:

- **Ego effects**: How an actor’s own covariate value influences their tendency to form or maintain ties (e.g., do smokers send more ties?).
- **Alter effects**: How the recipient’s covariate value affects the likelihood of receiving a tie (e.g., are smokers more likely to be chosen as friends?).
- **Similarity effects**: Whether actors prefer to form ties with others who are similar to themselves (e.g., are adolescents more likely to befriend others with similar smoking behavior?).

In Model 2, we treat sex as a fixed covariate and smoking behavior as a changing covariate observed at each wave. We include the following additional effects:

- **Ego effect of smoking**: Does an adolescent’s smoking status influence how many ties they send?
- **Alter effect of smoking**: Are smokers more likely to receive ties?
- **Similarity in smoking**: Are adolescents more likely to form ties with others who have similar smoking behavior?
- **Sex effects**: Do patterns of tie formation differ by sex?

```{r}
#| message: false
# Convert array into RSiena dependent object
siena_net <- sienaDependent(net_array)
```

Like before, this object represents the longitudinal network that serves as the
dependent variable in the SAOM. Different from before, we extract covariate
values from the network's node attributes. Sex (`sex.F`) is treated as a
time-invariant covariate, while smoking behavior (`familysmoking`) is treated as
time-varying across the three network waves.

Extract and recode sex to numeric, then convert to a covariate object.

```{r}
sex_char <- V(glasgow_g2)$sex.F
sex_num <- ifelse(sex_char == "F", 1, ifelse(sex_char == "M", 0, NA))
sex_cov <- coCovar(sex_num)
```

Extract smoking behavior which is assumed to change.

```{r}
c(smoke1, smoke2, smoke3) %<-%
  map(list(glasgow_g1, glasgow_g2, glasgow_g3),
      \(x) as.numeric(V(x)$tobacco))
smoking_array <- cbind(smoke1, smoke2, smoke3)
smoking_cov <- varCovar(smoking_array)
```

Combine network and covariates into Siena data object.

```{r}
data_mod2 <- sienaDataCreate(siena_net, sex_cov, smoking_cov)
```

Define effects to be included, starting with default set. Add structural, sex, and smoking effects.

```{r}
my_effects <- getEffects(data_mod2) %>%
  includeEffects(
    outdegree, recip, transTrip
  ) %>%
  includeEffects(
    egoX, altX, simX,
    interaction1 = "sex_cov"
  ) %>%
  includeEffects(
    egoX, altX, simX,
    interaction1 = "smoking_cov"
  )
```

Define estimation settings. and run model.

```{r}
#| output: false
#| cache: true
my_algorithm <- sienaAlgorithmCreate(projname = "saom_2", seed = 1108)

saom_2 <- siena07(
  my_algorithm,
  data = data_mod2,
  effects = my_effects,
  batch = T,
  useCluster = T, nbrNodes = 6
)
```

```{r}
saom_2
significance(saom_2)
```


### Convergence assessment

Individual and overall convergence t-ratios indicate good convergence.

### Rate parameters

Actors had about 11 opportunities to change between the first waves, and 9 chances to do so between the subsequent waves.

### Network dynamics

Again, reluctance to form many ties and tendency to reciprocity are indicated. There is also a preference for triadic closure. This represents sparsity, reciprocity and local clustering.

For covariate related effects:

Sex alter and sex ego are not statistically significant, indicating no systematic preference or avoidance of alters based on sex, nor any difference between males and females. However, there is a strong tendency toward homophily.

The coefficient for sex similarity is 0.9200 (SE = 0.1006). The coefficient is centered around the mean value.


```{r}
mean(sex_cov)
```
i.e., $\bar{v} = 0.434$. Thus, approximately 43.4% of actors in this dataset are
girls. In `RSiena`, covariates are mean-centered by default, ensuring that
estimated effects are interpreted relative to the average actor.

With 0 for boys and 1 for girls, the centered values are 

```{r}
(sex_center <- tibble(
  sex = c("boy", "girl"),
  value = c((0 - mean(sex_cov)), (1 - mean(sex_cov))),
))
```


Because covariates are mean-centered, the contribution of a potential tie
$y_{ij}$ to the objective function depends on these centered values. The
covariate-related part of the objective function for actor $i$, when evaluating
a candidate network $Y^{(a)}$, can be written as:

$$
f_i(\theta, Y^{(a)}) =
\theta_{\text{ego}} (v_i - \bar{v}) \sum_{j} y^{(a)}_{ij}
\;+\;
\theta_{\text{alter}} \sum_{j} y^{(a)}_{ij} (v_j - \bar{v})
\;+\;
\theta_{\text{sim}} \sum_{j} y^{(a)}_{ij} \, I(v_i = v_j)
$$

where:

- $v_i$ and $v_j$ are the covariate values (sex) for actors $i$ and $j$,
  respectively (coded as 0 = boy, 1 = girl),
- $\bar{v} = 0.434$ is the sample mean of the covariate,
- $y^{(a)}_{ij}$ indicates the presence of a tie from $i$ to $j$ in the
  candidate network $Y^{(a)}$,
- $I(v_i = v_j)$ is an indicator function equal to 1 if actors $i$ and $j$ share
  the same sex, and 0 otherwise.


```{r}
center_sex <- function(mod, var) {
    c(
    round(mod$theta[var] * 
            sex_center[sex_center$sex == "boy", ]$value, 3),
    round(mod$theta[var] * 
            sex_center[sex_center$sex == "girl", ]$value, 3)
  )
}

c(ego_sex, alter_sex) %<-%
  map(c(5, 4),
      \(x) center_sex(saom_2, x))
sim_sex <- c(round(saom_2$theta[6], 3), 0)
```

Total contribution to the objective function is then

```{r}
contribution <- function(ego_alt, ego_term, alt_term, sim_term) {
  length = length(ego_alt)
  tibble(
    ego = rep(ego_alt, each = length),
    alter = rep(ego_alt, length),
    ego_term = rep(ego_term, each = length),
    alter_term = rep(alt_term, length),
    similarity_term = sim_term
  ) %>%
    mutate(
      total_contribution = rowSums(across(where(is.numeric)))
    )
}

sim_term <- c(sim_sex, rev(sim_sex))

contribution(c("Male", "Female"), ego_sex, alter_sex, sim_term)
```

These values show that same-sex ties, especially between girls, are more likely
due to the strong similarity effect. Cross-sex ties are less likely, receiving
lower utility in the objective function.

The smoking variable is ordinal with three categories: 1 = no smoking, 2 = occasional smoking, and 3 = regular smoking. The covariate is centered by subtracting its mean. In this case, the mean value is computed in R as:

Smoking is encoded 1 for non-smoker, 2 for occasional smoker, 3 for regular smoker


```{r}
mean(smoking_cov)
```

```{r}
(smoke_center <- tibble(
  smokes = c("none", "occasional", "regular"),
  value = c((1 - mean(smoking_cov)), (2 - mean(smoking_cov)),
            (3 - mean(smoking_cov))),
))
```


The contribution of a potential tie $y_{ij}(t)$ (from actor $i$ to actor $j$) to
the objective function is given by:

$$
\theta_{\text{ego}}(v_i - \bar{v}) 
+ \theta_{\text{alter}}(v_j - \bar{v}) 
+ \theta_{\text{sim}} \left(1 - \frac{|v_i - v_j|}{R_v}\right)
$$

where $R_v$ denotes the range of the covariate $v$ (here, $R_v = 3 - 1 = 2$).
Substituting the estimated parameter values from the model:

$$
0.077(v_i - 1.377) + 0.109(v_j - 1.377) 
+ 0.380 \left(1 - \frac{|v_i - v_j|}{2} \right)
$$

This expression combines:

- an **ego effect**, capturing how smoking behavior influences how many ties
  actor $i$ tends to send,
- an **alter effect**, capturing how smoking behavior affects how attractive
  actor $j$ is as a recipient of ties,
- and a **similarity effect**, which rewards ties between actors with similar
  smoking levels.

```{r}

center_smoke <- function(var) {
    c(
    round(saom_2$theta[var] * 
            smoke_center[smoke_center$smokes == "none", ]$value, 3),
    round(saom_2$theta[var] * 
            smoke_center[smoke_center$smokes == "occasional", ]$value, 3),
    round(saom_2$theta[var] * 
            smoke_center[smoke_center$smokes == "regular", ]$value, 3)
  )
}

c(ego_smoke, alter_smoke) %<-%
  map(c(8, 7), center_smoke)
```

The similarity term is partial for none/occasional and occasional/regular.

```{r}
(sim_smoke <- 
   c(
     round(saom_2$theta[9], 3),
     round(saom_2$theta[9] / 2, 3),
     0
   ))
```

```{r}
sim_term <- c(sim_smoke, sim_smoke[2],
                      sim_smoke[1], sim_smoke[2],
                      rev(sim_smoke))

contribution(c("None", "Occasional", "Regular"), 
             ego_smoke, alter_smoke, sim_term)
```

@tbl-smokeeff illustrates how these smoking-related covariate effects contribute
to the objective function. Each cell reports the total contribution associated
with a potential tie from ego $i$ to alter $j$, given their respective smoking
levels.

Actors with similar smoking behavior are more likely to form ties, especially regular-to-regular smokers, indicating a strong homophily pattern. The ego and alter terms slightly amplify or reduce this tendency depending on individual smoking levels. Actors are increasingly likely to form ties the more they smoke.

## Model 3: Network and Behavioral Co-Evolution

We now estimate a co-evolution model in which the friendship network and smoking behavior are treated as jointly dependent processes. The network process models changes in friendship ties, while the behavioral process models changes in actors’ smoking behavior over time.

In this specification, smoking enters the model in two ways. First, smoking is used in the network objective function to test selection effects: whether actors form or maintain ties based on smoking behavior. Second, smoking is modeled as a behavioral dependent variable to test influence effects: whether actors adjust their smoking behavior in response to their friends.

Dependent variables

```{r}
siena_net <- sienaDependent(net_array)
smoking_beh <- sienaDependent(smoking_array, type = "behavior")
```

Covariate and datra object

```{r}
sex_cov <- coCovar(sex_num)
data_coev <- sienaDataCreate(
  siena_net, smoking_beh, sex_cov
)
```

Effects

Network Dynamics

```{r}
eff <- getEffects(data_coev) %>% 
  includeEffects(outdegree, recip, transTrip, name = "siena_net") %>% 
  includeEffects(egoX, altX, simX, 
                 name = "siena_net", interaction1 = "sex_cov") %>% 
  includeEffects(egoX, altX, simX, name = "siena_net",
                 interaction1 = "smoking_beh")
```

Behavior dynamics - shape, position, social influence

```{r}
eff <- eff %>% 
  includeEffects(linear, quad, name = "smoking_beh") %>% 
  includeEffects(indeg, outdeg, name = "smoking_beh",
                 interaction1 = "siena_net") %>% 
  includeEffects(avAlt, name = "smoking_beh", interaction1 = "siena_net")
```


We define the estimation settings using `sienaAlgorithmCreate()` and estimate
the co-evolution model.



```{r}
#| message: false
#| cache: true
#| output: false
alg <- sienaAlgorithmCreate(projname = "saom_coev", seed = 1108)

saom_coev <- siena07(
  alg,
  data = data_coev,
  effects = eff,
  batch = TRUE
)
```

Finally, we display the model results.

```{r}
saom_coev
significance(saom_coev)
```


The co-evolution model shows the following key patterns:

- **Network structure**
  - Strong negative outdegree $\Rightarrow$ sparse network
  - Strong positive reciprocity $\Rightarrow$ mutual ties are strongly favored
  - Positive transitivity $\Rightarrow$ evidence of triadic closure
- **Sex effects**
  - Strong positive similarity $\Rightarrow$ clear same-sex homophily
  - Ego and alter effects not significant $\Rightarrow$ no strong differences in
    tie activity by sex
- **Smoking and network selection**
  - Positive similarity effect $\Rightarrow$ friendships are formed between
    actors with similar smoking behavior (selection)
  - Ego and alter effects not significant $\Rightarrow$ smoking does not affect
    sending/receiving ties overall
- **Behavioral dynamics (smoking)**
  - Negative linear + positive quadratic shape $\Rightarrow$ tendency toward
    moderate smoking levels
  - Average alter effect positive and significant $\Rightarrow$ evidence of
    social influence (actors adjust smoking toward friends)
  - Indegree and outdegree effects not significant $\Rightarrow$ network
    position not strongly related to smoking

Overall, the results suggest that network evolution is primarily driven by
reciprocity, triadic closure, and homophily. In addition, there is evidence for
both selection and influence processes in the co-evolution of networks and
behavior: actors tend to form ties with others who have similar smoking
behavior, and they also adjust their behavior over time to become more similar
to their friends.


