---
title: "Descriptive Network Analysis"
format: gfm
---

[Source](https://schochastics.github.io/R4SNA/tidy/tidygraph-descriptive.html)

```{r}
options(paged.print = FALSE)
```

```{r}
#| message: false

libraries <- list(
  "tidygraph", "networkdata", "tidyverse", 
  "ggraph", "zeallot", "graphlayouts"
)
invisible(lapply(libraries, library, character.only = TRUE))
```

```{r}
flo_tidy <- as_tbl_graph(flo_marriage)
```


# Centrality

Centrality indices live in the centrality_*() function family. The family covers everything implemented in igraph — degree, closeness, betweenness, eigenvector, PageRank, Katz, hub and authority scores — together with the richer catalogue exposed by the netrankr package. 

```{r}
#| label: fig-centrality-example
#| fig-cap: "Florentine marriage network with node size scaled to degree centrality and fill scaled to betweenness centrality."

flo_tidy %>% 
  activate("nodes") %>% 
  mutate(
    degree = centrality_degree(),
    betweenness = centrality_betweenness()
  ) %>% 
  ggraph("stress", bbox = 10) +
  geom_edge_link0(edge_color = "black") +
  geom_node_point(aes(size = degree, fill = betweenness), shape = 21) +
  geom_node_text(aes(label = name)) +
  scale_fill_gradient(low = "#104E8B", high = "#CD2626") +
  scale_size(range = c(4, 10)) +
  theme_graph()
```

# Clustering

Community detection algorithms are exposed through the parallel `group_*()`
family: `group_louvain`, `group_infomap`, `group_edge_betweenness`,
`group_walktrap`, `group_leading_eigen`, `group_fast_greedy`,
`group_label_prop`, `group_spinglass`, `group_components`,
`group_biconnected_component`, and a few more. Each returns an integer
membership vector the length of the node table, which means the idiomatic usage
is `mutate(community = as.factor(group_*()))` --- coerce to factor so the result
becomes a natural input to a `fill` or `color` aesthetic.

```{r}
#| label: fig-clustering-louvain
#| fig-cap: "Louvain community detection applied to a graph with planted community structure generated by `play_islands()`. Node fill encodes the recovered community membership."

play_islands(4, 12, 0.8, 4) %>% 
  mutate(community = as.factor(group_louvain())) %>% 
  ggraph("stress") +
  geom_edge_link0() +
  geom_node_point(aes(fill = community), shape = 21, size = 6) +
  theme_graph()
```

Color edges by community.

```{r}
#| label: fig-clustering-bridges
#| fig-cap: "Same graph as in @fig-clustering-louvain, with edges additionally colored to distinguish intra-community ties from bridges between communities."

play_islands(4, 12, 0.8, 4) %>% 
  mutate(community = as.factor(group_louvain())) %>% 
  activate("edges") %>% 
  mutate(
    community = as.factor(ifelse(
      .N()$community[from] == .N()$community[to],
      .N()$community[from], 5
    ))
  ) %>% 
  ggraph("stress") +
  geom_edge_link0(aes(edge_colour = community), show.legend = F) +
  geom_node_point(aes(fill = community), shape = 21, size = 6) +
  scale_fill_brewer(palette = "Set3") +
  scale_edge_color_brewer(palette = "Set3") +
  theme_graph(background = "grey88")
```

# Other node or edge level functions

Beyond centrality and grouping, tidygraph harmonises a long list of node- and
edge-level measures behind two mirror-image function families: `node_*()`
returns a vector of length `vcount()`, and `edge_*()` returns a vector of length
`ecount()`. Which family you reach for is determined by which table you have
activated. The families cover distance (`node_distance_to`,
`node_distance_from`, `node_eccentricity`), local structure (`node_triangles`,
`node_coreness`, `node_local_transitivity`), and tests on edges (`edge_is_loop`,
`edge_is_multiple`, `edge_is_mutual`, `edge_is_incident`), among many others. As
with the centrality and group families, typing `node_` or `edge_` at the prompt
and pressing Tab gives the full list.

```{r}
#| label: fig-distance-medici
#| fig-cap: "Florentine marriage network with node fill encoding the geodesic distance to the Medici family and edge color flagging ties incident to the Medici."

flo_tidy %>% 
  activate("nodes") %>% 
  mutate(dist2Medici = node_distance_to(nodes = 9)) %>% 
  activate("edges") %>% 
  mutate(edge2Medici = edge_is_incident(9)) %>% 
  ggraph("stress") +
  geom_edge_link0(aes(edge_color = edge2Medici)) +
  geom_node_point(aes(fill = dist2Medici), size = 9, shape = 21) +
  theme_graph()
```

The node id 9 refers to the Medici, which happens to be the ninth row of the node table. Passing the index directly is the quickest way when you already know the position; for a name-based lookup you can equally write nodes = which(.N()$name == "Medici").















